---
title: "Categorical Data in R"
author: "Arvind Venkatadri"
date: 2023/16/01
lastmod: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: tango
    toc_float: TRUE
    toc_depth: 3
    df_print: paged
    number_sections: TRUE
    code_folding: show
    code_download: TRUE
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

Let us play with some categorical ( or predominantly categorical )
datasets in R and see how we can analyze and plot them.

First we will learn how to make Contingency Tables with any of the three forms. This will be useful in arriving at a common form of Table to go with plotting.

Then we will use `vcd`, `mosaic`, `ggplot` and `ggpubr` to make several plots for Categorical Datasets. 


```{r setup, include=TRUE}
knitr::opts_chunk$set(message = FALSE, echo = TRUE)
library(tidyverse)
library(vcd) #COntingency Tables and Plots
library(vcdExtra) # Datasets
library(sjPlot) # Likert Plots
library(mosaic) # Data Analysis and Plots
library(ggmosaic) # Mosaic Plots
library(ggpubr) # Balloon Plots

#install.packages("openintro")
library(openintro) # More Datasets

```


# Creating Contingency Tables {.tabset .tabset-pills}

Most plots for Categorical Data ( as we shall see ) require that the data be converted into a *Contingency Table*; even Statistical tests for Proportions ( the $ X^2 $ test ) need Contingency Tables. The *Frequency Table* we encountered earlier is very close to being a full-fledged Contingency Table. ( needs to add row and column margin counts )

In this section we understand how to make Contingency Tables from each of the three forms.

## Using base R

```{r}
Arthritis

#One Way Table ( one variable )
table(Arthritis$Treatment)
table(Arthritis$Treatment) %>% prop.table()
table(Arthritis$Treatment) %>% addmargins()
```


```{r}

# Two-Way Table ( two variables )
table(Arthritis$Treatment, Arthritis$Improved) %>% prop.table() 
table(Arthritis$Treatment, Arthritis$Improved) %>% addmargins() # Contingency Table!!

```

We can use **table()** and **xtabs()** to generate multi-dimensional tables too ( More than 2D ) These will be printed out as a series of 2D tables, one for each value of the "third" parameter. 

We can also **ftable()** to print these out in an attractive manner.

```{r}

mytable <- table(Arthritis$Treatment, Arthritis$Sex, Arthritis$Improved)
mytable
ftable(mytable)
ftable(mytable) %>% addmargins()

```


## Using the `vcd` package

The `vcd` ( Visualize Categorical Data ) package by Michael Friendly has a convenient function to create Contingency Tables: `structable()`; this function produces a ‘flat’ representation of a high-dimensional contingency table constructed by recursive splits (similar to the construction of mosaic charts/graphs).

The arguments of `structable` are:  

- a formula ($y + p \sim x + z $ ) which shows which variables are to be included as *columns* and *rows* respectively on a table  
- a `data` argument, which can indicate a `data frame`


```{r}
vcd::structable(data = Arthritis, Treatment ~ Improved)
vcd::structable(data = Arthritis, Treatment ~ Improved + Sex)


# HairEyeColor is in multiple table form
# structable flattens these into one, as for a mosaic chart
HairEyeColor
vcd::structable(HairEyeColor)

## UCBAdmissions is already in Frequency Form i.e. a Contingency Table
#`structable` tends to render flat tables, of the kind that can be thought of as a "text representation" of the `vcd::mosaic` plot
UCBAdmissions
vcd::structable(UCBAdmissions)

```


## Using the `mosaic` package

I think this is the simplest and most elegant way of obtaining Contingency Tables:

```{r}
# One Way Table
tally( ~ substance, data = HELPrct)

# Two Way Tables
tally( ~ substance + sex , data = HELPrct)
tally( ~ substance | sex , data = HELPrct)

tally( sex ~ substance, data = HELPrct)
tally(~ sex |substance, data = HELPrct)

```


```{r,highlight=TRUE}

# Adding Margins
tally( ~ substance + sex , data = HELPrct, format = 'count', margins = TRUE) # Ta Da!
tally( ~ substance + sex , data = HELPrct, format = 'percent', margins = TRUE)

```

### Using the `tidyverse`


```{r}
diamonds %>% group_by(cut, clarity) %>% dplyr::summarise( count = n())

# We need to pivot this "wide" to obtain a Contingency Table

diamonds %>% 
  group_by(cut, clarity) %>% 
  dplyr::summarise( count = n()) %>% 
  pivot_wider(id_cols = cut, names_from = clarity, values_from = count) %>% 
  
  # Now add the row and column totals using the `janitor` package
  janitor::adorn_totals(where = c("row", "col")) %>%
  
  # Recovert to tibble since janitor gives a "tabyl" format ( which can be useful )
  as_tibble()

```



Now that we have Contingency Tables, we can plot these:

# Plotting Categorical Data

## The `titanic` dataset

```{r}
data("titanic")
titanic

```


### `titanic` Bar Plots

```{r titanic-bar-plot}
# Use dplyr and ggplot




```




### `titanic` Mosaic Plot

```{r titanic-mosaic-plot}
# Try the mosaic package and the ggmosaic package




```

### `titanic` Balloon Plot

```{r titanic-balloon-plot}
# use ggpubr




```


## The `hippocorpus` dataset from Kaggle

```{r, echo=FALSE,message=FALSE}
library(downloadthis)
hippo <- read.csv("data/hippoCorpusV2.csv")
download_this(hippo,
    #path = "data/hippoCorpusV2.csv",
    output_name = "hippocorpus",
    output_extension = ".csv",
    button_label = "Download data as csv",
    button_type = "info",
    has_icon = TRUE,
    icon = "fa fa-save"
  )

```

This is a dataset from
[Kaggle](https://www.kaggle.com/datasets/saurabhshahane/hippocorpus?select=hippoCorpusV2.csv)
and is based on Reference 2.

> Hippocorpus is dataset of 6854 English diary like short stories about
> recalled and imagined events. Using a crowdsourcing framework the
> respective owners of this datasets collected recalled stories and
> summaries from workers, then provided these collected summaries to
> other workers who write imagined stories. Months later dataset
> creators collected a retold version of the recalled stories from the
> subset of recalled authors. Dataset contains author demographics (age,
> gender, race), their openness to experience, as well as some variables
> regarding the author's relationship to the event (e.g., how personal
> the event is, how often they tell its story, etc.)

Apart from metadata pertaining to each respondent, there are 4 *Likert
Scale* variables:

-   `distracted`: How distracted were you while writing your story?
    (5-point Likert)
-   `draining`: How taxing/draining was writing for you emotionally?
    (5-point Likert)
-   `frequency`: How often do you think about or talk about this event?
    (5-point Likert)
-   `importance`: How impactful, important, or personal is this
    story/this event to you? (5-point Likert). Plot these using the
    package `sjPlot`. Can you also try a `ggplot`?

```{r hippocorpus-likert}


```


## A dataset from the `vcdExtra` package

Pick one of the fairly large Categorical datasets that are built into `vcdExtra`: type `data(package = "vcdExtra")` in your Console.

Create:  
- Contingency Table
- A Bar Plot
- A Mosaic Plot
- A Balloon Plot


# Conclusion

Write a few comments on the data and visualizations. Did they convey a story of sorts?


# References

1.  A detailed analysis of the NHANES dataset,
    <https://awagaman.people.amherst.edu/stat230/Stat230CodeCompilationExampleCodeUsingNHANES.pdf>

2.  Maarten Sap, Eric Horvitz, Yejin Choi, Noah A. Smith, and James
    Pennebaker (2020) *Recollection versus Imagination: Exploring Human
    Memory and Cognition via Neural Language Models.* ACL.
