---
title: "\U0001F554 Time Series"
author: Arvind Venkatadri
date: "`r Sys.Date()`"
tags:
  - CandleStick Graphs
  - Heatmap Graphs (over time)
  - Line Graphs
  - Time Series
subtitle: Time Series
weight: 80
summary: Trends, Seasons, and Changes over time
image:
  caption: Photo by https://twitter.com/culturaltutor/status/1605415692169347073
  focal_point: Smart
slides: ""
url_code: ""
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(magrittr)
library(lubridate)  # Deal with dates

library(mosaic)

library(tsibble) # Create and manipulate Time Series in R
library(tsibbledata) # Tsibble Time Series Datasets
library(fable) # Modelling Time Series in R
library(fabletools)
library(feasts) # Feature Extraction and Statistics for Time Series.
library(timetk)  # Convert data frames to time series-specific objects
library(forecast)  # Make forecasts and decompose time series

# devtools::install_github("FinYang/tsdl")
library(tsdl) # Time Series Data Library from Rob Hyndman


```


## Introduction

Any metric that is measured over regular time intervals forms a **time
series**. Analysis of Time Series is commercially important because of
industrial need and relevance, especially with respect to Forecasting
(Weather data, sports scores, population growth figures, stock prices,
demand, sales, supply...). In the graph shown below is the temperature 
over times in two US cities:

<iframe width="100%" height="514" frameborder="0"
  src="https://observablehq.com/embed/f274bd3d517cf495?cells=LineChart"></iframe>
  

A time series can be broken down to its components so as to
systematically understand, analyze, model and forecast it. We have to
begin by answering fundamental questions such as:

1.  What are the types of time series?  
1.  How to decompose it? How to extract a level, a trend, and seasonal components from a time series?  
1.  What is auto correlation etc.  
1.  What is a stationary time series?  
1.  And, how do you plot time series?  


# Introduction to Time Series:  Data Formats

There are multiple formats for time series data.

- Tibble format: the simplest is of course the standard 
tibble/ dataframe, with a `time` variable to indicate that the other 
variables vary with time.

- The `ts` format: The `stats::ts()` function will convert a **numeric 
vector** into an R time series `ts` object. 

- The modern `tsibble` format: this is a new format for time series 
analysis, and is used by the [tidyverts](https://tidyverts.org) set of packages.



- The base `ts` object is used by established packages `forecast`
- The standard tibble object is used by `timetk` & `modeltime`
- The special `tsibble` object ("time series tibble") is used by `fable`, `feasts` and others from the `tidyverts` set of packages


## Creating and Plotting Time Series

In this first example, we will use simple `ts` data first, and then do 
another with `tsibble` format, and then a third example with a `tibble` that we can plot as is and do more after conversion to `tsibble` format. 

### `ts` format data

There are a few datasets in base R that are in `ts` format already. 

```{r ts-data}

AirPassengers
str(AirPassengers)

```

This can be easily plotted using base R:

```{r}

plot(AirPassengers)

```

One can see that there is an upward *trend* and also *seasonal variations* that also increase over time.

Let us take data that is "time oriented" but not in `ts` format: the 
syntax of `ts()` is:

    Syntax:  objectName <- ts(data, start, end, frequency)

    where,

        `data`: represents the data vector
        `start`: represents the first observation in time series
        `end`: represents the last observation in time series
        `frequency`: represents number of observations per unit time. For 
        example 1=annual, 4=quarterly, 12=monthly, etc.

We will pick simple numerical vector data variable from `trees`:

```{r trees-ts}

trees

# Choosing the `height` variable
trees_ts <- ts(trees$Height, 
               frequency = 1, # No reason to believe otherwise
               start = 1980)  # Arbitrarily picked "1980" !
plot(trees_ts)

```
( Note that this example is just for demonstration: tree heights do not decrease over time!!) 

## `tsibble` data

The package `tsibbledata` contains several ready made `tsibble` format data. Run `data(package = "tsibbledata")` to find out about these.
Let us try `PBS` which is a dataset containing Monthly Medicare prescription data in Australia.

```{r PBS}

data("PBS")

```
This is a large dataset, with 1M observations, for 336 combinations of `key` variables. Data appears to be *monthly*. Note that there is **more than one** quantitative variable, which one would not be able to support in the `ts` format. 

There are multiple **Quantitative variables** ( `Scripts` and `Cost`). The **Qualitative Variables** are described below. (Type `help("PBS")` in your Console)

>The data is disaggregated using four keys:

>Concession:	Concessional scripts are given to pensioners, unemployed, dependents, and other card holders  
Type:	Co-payments are made until an individual's script expenditure hits a threshold ($290.00 for concession, $1141.80 otherwise). Safety net subsidies are provided to individuals exceeding this amount.  
ATC1:	Anatomical Therapeutic Chemical index (level 1)
ATC2:	Anatomical Therapeutic Chemical index (level 2)

Let us simply plot `Cost` over time:

```{r PBS-Cost}

PBS %>% ggplot(aes(x = Month, y = Cost)) + 
  geom_point() + 
  geom_line()

```

This basic plot is quite messy. We ought to use `dplyr` to filter the data using some combination of the Qualitative variables( 336 combinations!). Let us try that now:

```{r}

PBS %>% count(ATC1, ATC2, Concession, Type)

```
We have 336 combinations of **Qualitative** variables, each containing 204 observations: so let us filter for a few such combinations and plot:

```{r}

PBS %>% dplyr::filter(Concession == "General", 
                      ATC1 == "A",
                      ATC2 == "A10") %>% 
  ggplot(aes(x = Month, y = Cost, colour = Type)) + 
  geom_line() + 
  geom_point()

```

As can be seen, very different time patterns based on the two `Type`s of
payment methods.  Strongly seasonal for both, with seasonal variation increasing over the years, but there is an upward *trend* with the Co-payments method of payment. 


## `tibble` data


Let us read and inspect in the US births data from 2000 to 2014. Download this data by clicking on the icon below, and saving the downloaded file in a sub-folder called `data` inside your project:

```{r echo=FALSE}
# Don't install this if you are a student !! ;-D
library(downloadthis)
download_file(
  path = "data/US_births_2000-2014_SSA.csv",
  output_extension = ".csv",
  output_name = "US_births_2000-2014_SSA",
  button_label = "Download US Births Data",
  button_type = "info",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

```

Read this data in:

```{r read-in-data,echo=FALSE,message=FALSE}

births_2000_2014 <- read_csv("data/US_births_2000-2014_SSA.csv")

# # This data goes up to 2003, but the previous data starts at 2000, so we'll
# # remove 2000-2003 from here
# births_1994_1999 <- read_csv("data/US_births_1994-2003_CDC_NCHS.csv") %>% 
#   filter(year < 2000)

glimpse(births_2000_2014)

```

So there are several numerical variables for `year`, `month`, and `day_of_month`, `day_of_week`, and of course the `births` on a daily basis. We will create a `date` column with these separate ones above, and then plot the births, say for the month of March, in each year:


```{r, echo=FALSE}

births_tsibble <- births_2000_2014 %>% 
  mutate(date = lubridate::make_date(year = year,
                                     month = month,
                                     day = date_of_month)) %>%
  
  # drop year month day since we have date
  select(date, births, date_of_month,day_of_week) %>% 
  tsibble::as_tsibble(index = date # Time Variable
                      )

births_tsibble

births_tsibble %>%
  tsibble::index_by(month_index = ~ yearmonth(.)) %>% 
  # Grouping by month_index
  # NOT THE SAME AS `group_by(month)` !!!
  # if we had kept the original `month` variable !!!
  
  
  # Absolutely need this to take time averages
  summarise(mean_births = mean(births, na.rm = TRUE)) %>% 
  
  ggplot(., aes(x = month_index, y = mean_births)) + 
  geom_point() +
  geom_line()

```

Hmm...can we try to plot box plots over time (Candle-Stick Plots)? Over month / quarter or year?



```{r,message=FALSE}

# Monthly box plots
births_tsibble %>%
  index_by(month_index = ~ yearmonth(.)) %>% # 180 months over 15 years
  # No need to summarise, since we want boxplots per year / month
  ggplot(., aes(y = births, x = date, 
                group =  month_index)) + # plot the groups
  
  geom_boxplot(aes(fill = month_index))      # 180 plots!!  
  
# Quarterly boxplots
births_tsibble %>%
  index_by(qrtr_index = ~ yearquarter(.)) %>% # 60 quarters over 15 years
  # No need to summarise, since we want boxplots per year / month
  ggplot(., aes(y = births, x = date, 
                group = qrtr_index)) +
  
  geom_boxplot(aes(fill = qrtr_index))        # 60 plots!!


# Yearwise boxplots
births_tsibble %>% 
  index_by(year_index = ~ lubridate::year(.)) %>% # 15 years, 15 groups
    # No need to summarise, since we want boxplots per year / month

  ggplot(., aes(y = births, 
                x = date, 
                group = year_index)) + # plot the groups
  
  geom_boxplot(aes(fill = year_index)) +           # 15 plots
  scale_fill_distiller(palette = "Spectral")

```

Although the graphs are very busy, they do reveal seasonality trends at different periods.


## Seasons, Trends, Cycles, and Random Changes

> Trend
A trend exists when there is a long-term increase or decrease in the data. It does not have to be linear. Sometimes we will refer to a trend as “changing direction”, when it might go from an increasing trend to a decreasing trend. 

>Seasonal
A seasonal pattern occurs when a time series is affected by seasonal factors such as the time of the year or the day of the week. Seasonality is always of a fixed and known period. The monthly sales of drugs ( with the PBD data )  shows seasonality which is induced partly by the change in the cost of the drugs at the end of the calendar year.

>Cyclic
A cycle occurs when the data exhibit rises and falls that are not of a fixed frequency. These fluctuations are usually due to economic conditions, and are often related to the “business cycle”. The duration of these fluctuations is usually at least 2 years.

Let us try to find and plot these patterns in Time Series.


```{r,STL-model-yearly}

births_STL_yearly <- births_tsibble %>% 
  fabletools::model(STL(births ~ season(period = "year")))

fabletools::components(births_STL_yearly)
feasts::autoplot(components(births_STL_yearly))

```


How about a heatmap? We can cook up a categorical variable based on the number of births (low, fine, high) and use that to create a heatmap:


```{r,echo=FALSE}


library(ggformula)

births_2000_2014 %>%  
  mutate(birthrate = case_when(births >=10000 ~ "high", births <= 8000 ~ "low", TRUE ~ "fine")) %>% 
  gf_tile(data = ., year ~ month, fill = ~ birthrate, color = "black") %>%
  
  gf_theme(scale_x_time(breaks = 1:12, labels = c("Jan", "Feb", "Mar","Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))) %>% 
  
  gf_theme(theme_classic())

```



## A Workflow in R

Download the RMarkdown tutorial file by clicking the icon above and open
it in RStudio or rstudio.cloud.

```{r, echo=FALSE}
# Don't install this if you are a student !! ;-D
library(downloadthis)
download_file(
  path = "files/timeseries.Rmd",
  output_extension = ".Rmd",
  output_name = "r-timeseries",
  button_label = "Download RMarkdown Workflow",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

## Readings

1. [The Nuclear Threat---The Shadow Peace, part
    1](http://www.fallen.io/shadow-peace/1/)
    
1.  [11 Ways to Visualize Changes Over Time -- A
    Guide](https://flowingdata.com/2010/01/07/11-ways-to-visualize-changes-over-time-a-guide/)


1.  [What is seasonal adjustment and why is it
    used?](http://junkcharts.typepad.com/junk_charts/2010/11/what-is-seasonal-adjustment-and-why-is-it-used.html)
    
1.  [The start-at-zero
    rule](http://junkcharts.typepad.com/junk_charts/2005/09/the_startatzero.html)
1.  [Keeping one's appetite after touring the sausage
    factory](http://junkcharts.typepad.com/numbersruleyourworld/2011/02/keeping-ones-appetite-after-touring-the-sausage-factory.html)
1.  [How Common is Your Birthday? This Visualization Might Surprise
    You](http://thedailyviz.com/2016/09/17/how-common-is-your-birthday-dailyviz/)

1.  [The Fallen of World War II](http://www.fallen.io/ww2/)
2.  [Visualizing Statistical Mix Effects and Simpson's
    Paradox](https://static.googleusercontent.com/media/research.google.com/en//pubs/archive/42901.pdf)
3.  [How To Fix a Toilet (And Other Things We Couldn't Do Without
    Search)](http://how-to-fix-a-toilet.com/)


------------------------------------------------------------------------
