---
title: "Lab 03a: An Introduction to the dplyr package"
author: "Arvind Venkatadri"
date: 06/July/2021
output:
  html_document:
    theme: flatly
    toc: TRUE
    toc_float: TRUE
    toc_depth: 2
    number_sections: TRUE
    code_folding: hide
    code_download: TRUE
abstract: Part of the `R for Artists and Designers` course at the School of Foundation Studies, Srishti Manipal Institute of Art, Design, and Technology, Bangalore.

---
# Introduction

This RMarkdown document is part of the Generic Skills Component  (GSK) of the Course of the  Foundation Studies Programme at Srishti Manipal Institute of Art, Design, and Technology, Bangalore India. The material is based on *A Layered Grammar of Graphics* by Hadley Wickham. The course is meant for First Year students pursuing a Degree in Art and Design. 

The intent of this GSK part is to build Skill in coding in R, and also appreciate R as a way to metaphorically visualize information of various kinds, using predominantly geometric figures and structures.

All RMarkdown files combine code, text, web-images, and figures developed using code. Everything is text; code chunks are enclosed in **fences** (```)

# Goals for Lab 03a

- Understand the idea of "tidy" data
- Using "tidy data" and the "tidyverse" way of programming in R allows to translate our thoughts readily into code. 
- Understand `dplyr` VERB functions to get to know and manipulate a dataset

# Pedagogical Note

The method followed will be based on
[PRIMM](https://blogs.kcl.ac.uk/cser/2017/09/01/primm-a-structured-approach-to-teaching-programming/):

-   **PREDICT** Inspect the code and guess at what the code might do,
    **write predictions**
-   **RUN** the code provided and check what happens
-   **INFER** what the `parameters` of the code do and **write comments to explain**. What bells and whistles can you see?
-   **MODIFY** the `parameters` code provided to understand the
    `options` available. **Write comments** to show what you have aimed for and achieved.
-   **MAKE** : take an idea/concept of your own, and graph it.
# Inspiration + data

We'll use data from the Star Wars series of movies. May the Force be with you!!

# Packages needed

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(tidyverse)
set.seed(1014)
```

When working with data you must:

* Figure out what you want to do.

* Describe those tasks in the form of a computer program.

* Execute the program.

The `dplyr` package makes these steps fast and easy:

* By constraining your options, it helps you think about your data manipulation challenges.

* It provides simple **"verbs"**, functions that correspond to the most common data manipulation tasks, to help you translate your thoughts into code.

This document introduces you to dplyr's basic set of tools, and shows you how to apply them to data frames. `dplyr` also supports databases via the `dbplyr` package, once you've installed, read `vignette("dbplyr")` to learn more.

## Data: starwars

To explore the basic data manipulation verbs of `dplyr`, we'll use the dataset `starwars`. This dataset contains `r nrow(starwars)` characters and comes from the [Star Wars API](https://swapi.dev), and is documented in `?starwars` 

This means: type `?starwars` in the Console. Try.

```{r}
dim(starwars)
starwars
```

Note that `starwars` is a `tibble`, a modern reimagining of the data frame. It's particularly useful for large datasets because it only prints the first few rows. You can learn more about tibbles at <https://tibble.tidyverse.org>; in particular you can convert data frames to tibbles with `as_tibble()`.

Check your Environment Tab to inspect `starwars` in a separate tab.

## Single table verbs

`dplyr` aims to provide a function for each basic verb of data manipulation. These verbs can be organised into three categories based on the component of the dataset that they work with:

* Rows:
  * `filter()` chooses rows based on column values.
  * `slice()` chooses rows based on location.
  * `arrange()` changes the order of the rows.
  
* Columns:
  * `select()` changes whether or not a column is included.
  * `rename()` changes the name of columns.
  * `mutate()` changes the values of columns and creates new columns.
  * `relocate()` changes the order of the columns.

* Groups of rows:
  * `summarise()` collapses a group into a single row.

Think of the parallels from Microsoft Excel. 

### The pipe

All of the `dplyr` functions take a data frame (or `tibble`) as the first argument. Rather than forcing the user to either save intermediate objects or nest functions, dplyr provides the `%>%` operator from `magrittr`. `x %>% f(y)` turns into `f(x, y)` so the result from one step is then "piped" into the next step. You can use the pipe to rewrite multiple operations that you can read left-to-right, top-to-bottom (**reading the pipe operator as "then"**). 

### Filter rows with `filter()`

`filter()` allows you to select a subset of rows in a data frame. Like all single verbs, the first argument is the tibble (or data frame). The second and subsequent arguments refer to variables within that data frame, selecting rows where the expression is `TRUE`.

For example, we can select all character with light skin color and brown eyes with:

Note the double equal to sign (==) below! Equivalent to MS Excel Data -> Filter

```{r}
starwars %>% filter(skin_color == "light", eye_color == "brown")
```


### Arrange rows with `arrange()`

`arrange()` works similarly to `filter()` except that instead of filtering or selecting rows, it **reorders** them. It takes a data frame, and a set of column names (or more complicated expressions) to order by. If you provide more than one column name, each additional column will be used to break ties in the values of preceding columns:

```{r}
starwars %>% arrange(height, mass)
```

Use `desc()` to order a column in descending order:

```{r}
starwars %>% arrange(desc(height))
```

###  Choose rows using their position with `slice()`

`slice()` lets you index rows by their (integer) locations. It allows you to select, remove, and duplicate rows. 

> This is an important step in Prediction, Modelling and Machine Learning. 

We can get characters from row numbers 5 through 10.
```{r}
starwars %>% slice(5:10)
```

It is accompanied by a number of helpers for common use cases:

* `slice_head()` and `slice_tail()` select the first or last rows.

```{r}
starwars %>% slice_head(n = 3)
```

* `slice_sample()` randomly selects rows. Use the option prop to choose a certain proportion of the cases.

```{r}
starwars %>% slice_sample(n = 5)
starwars %>% slice_sample(prop = 0.1)
```

Use `replace = TRUE` to perform a bootstrap sample. If needed, you can weight the sample with the `weight` argument.

        Note : ` Bootstrap samples` are a special statistical sampling method. Counterintuitive perhaps, since you sample **with replacement**. Should remind you of your high school Permutation and Combination class, with all those urn models and so on. If you remember. 

* `slice_min()` and `slice_max()` select rows with highest or lowest values of a variable. Note that we first must choose  only the values which are not NA.

```{r}
starwars %>%
  filter(!is.na(height)) %>%
  slice_min(height, n = 3)
```

### Select columns with `select()`

Often you work with large datasets with many columns but only a few are actually of interest to you. `select()` allows you to rapidly zoom in on a useful subset using operations that usually only work on numeric variable positions:

```{r}
# Select columns by name
starwars %>% select(hair_color, skin_color, eye_color)
# Select all columns between hair_color and eye_color (inclusive)
starwars %>% select(hair_color:eye_color)
# Select all columns except those from hair_color to eye_color (inclusive)
starwars %>% select(!(hair_color:eye_color))
# Select all columns ending with color
starwars %>% select(ends_with("color"))
```

There are a number of helper functions you can use within `select()`, like `starts_with()`, `ends_with()`, `matches()` and `contains()`. These let you quickly match larger blocks of variables that meet some criterion. See `?select` for more details.

You can even rename variables with `select()` by using named arguments:

```{r}
starwars %>% select(home_world = homeworld)
```

But because `select()` drops all the variables not explicitly mentioned, it's not that useful. Instead, use `rename()`:

```{r}
starwars %>% rename(home_world = homeworld)
```

### Add new columns with `mutate()`

Besides selecting sets of existing columns, it's often useful to add new columns that are functions of existing columns. This is the job of `mutate()`:

```{r}
starwars %>% mutate(height_m = height / 100)
```

We can't see the height in meters we just calculated, but we can fix that using a select command.

```{r}
starwars %>%
  mutate(height_m = height / 100) %>%
  select(height_m, height, everything())
```

`dplyr::mutate()` is similar to the base `transform()`, but allows you to refer to columns that you've just created:

```{r}
starwars %>%
  mutate(
    height_m = height / 100,
    BMI = mass / (height_m^2)
  ) %>%
  select(BMI, everything())
```

If you only want to keep the new variables, use `transmute()`:

```{r}
starwars %>%
  transmute(
    height_m = height / 100,
    BMI = mass / (height_m^2)
  )
```

### Change column order with `relocate()`

Use a similar syntax as `select()` to move blocks of columns at once

```{r}
starwars %>% relocate(sex:homeworld, .before = height)
```


### Summarise values with `summarise()`

The last verb is `summarise()`. It collapses a data frame to a single row.

```{r}
starwars %>% summarise(mean_height = mean(height, na.rm = TRUE))
```

It's not that useful until we learn the `group_by()` verb below.


### Commonalities

You may have noticed that the syntax and function of all these verbs are very similar:

* The first argument is a data frame.

* The subsequent arguments describe what to do with the data frame. You can
  refer to columns in the data frame directly without using `$`.

* The result is a new data frame

Together these properties make it easy to chain together multiple simple steps to achieve a complex result. 

These five functions provide the basis of a language of data manipulation. At the most basic level, you can only alter a tidy data frame in five useful ways: you can reorder the rows (`arrange()`), pick observations and variables of interest (`filter()` and `select()`), add new variables that are functions of existing variables (`mutate()`), or collapse many values to a summary (`summarise()`). 

## Combining functions with `%>%` 

The dplyr API is functional in the sense that function calls don't have side-effects. dplyr provides the `%>%` operator from magrittr. `x %>% f(y)` turns into `f(x, y)` so you can use it to rewrite multiple operations that you can read left-to-right, top-to-bottom (reading the pipe operator as "then"):

```{r}
starwars %>%
  group_by(species, sex) %>%
  summarise(
    mean_height = mean(height, na.rm = TRUE),
    mean_mass = mean(mass, na.rm = TRUE)
  )
```

