---
author: "Arvind Venkatadri"
date: '`r format(Sys.Date())`'
title: Using htmlwidgets, crosstalk, and echarts4r
output:
  html_document:
    toc_float: true
    theme: flatly
    toc: true
    code_folding: show
    code_download: true
    highlight: tango
keywords: echarts4r
subject: htmlwidgets
abstract: Using htmlwidgets package and check out its capabilities, especially maps. Also want to use crosstalk to link different visualizations together
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Data manipulation and data sources
library(tidyverse)
library(tidygraph)
library(igraph)
library(palmerpenguins)
library(igraphdata)

# htmlwidget related libraries
library(htmlwidgets)
library(crosstalk)

# Widget Libraries
library(leaflet)
library(plotly)
library(DT)
library(echarts4r)
library(rgl)
library(networkD3)
library(threejs)
library(slickR)
# Linkable widgets in crosstalk - github repo only
#devtools::install_github("kent37/summarywidget")
library(summarywidget)

```

## References

1. <https://www.htmlwidgets.org/index.html>
2. The threejs package: three.js widgets for R <https://bwlewis.github.io/rthreejs/>


## `htmlwidgets` usage

`htmlwidgets` creates, well, widgets, that can visualize data in many ways. HTML widgets work just like R plots except they produce interactive web visualizations. These can be used in `RMarkdown`, in `flexdashboards`, and in `shiny` apps. 

All the possible widgets ( 127 of them on CRAN ) are listed here: <https://gallery.htmlwidgets.org/>

Some packages that offer widgets for use in `htmlwidgets`:  

1. `networkD3`: 
  - Force directed networks with `simpleNetwork` and `forceNetwork`
  - Sankey diagrams with `sankeyNetwork`
  - Radial networks with `radialNetwork`
  - Dendro networks with `dendroNetwork`

### Using `networkD3`

```{r networkD3}
library(networkD3)
data("karate")

# Make separate data frames for edges and nodes
# networkD3 needs indices starting from 1
karate_edges <- karate %>% 
  as_tbl_graph() %>% 
  tidygraph::activate(edges) %>% 
  as.data.frame() %>% 
  dplyr::mutate(source = from - 1, target = to - 1) %>% 
  rename(value = weight) %>% 
  select(source, target, value)
karate_edges %>% head() %>% knitr::kable()

karate_nodes <- karate %>% 
  as_tbl_graph() %>% 
  tidygraph::activate(nodes) %>% 
  as.data.frame() %>% 
  dplyr::mutate(group = as.character(Faction)) %>% 
  select(name, label, group, color)
karate_nodes %>% head() %>% knitr::kable()

simpleNetwork(karate_edges, charge = -50, 
              fontSize = 12, zoom = FALSE, fontFamily = "serif")

```

```{r forceNetwork}
forceNetwork(Links = karate_edges, 
             Nodes = karate_nodes, 
             Value = "value", # width of edges, dbl
             NodeID = "name", # chr
             Group = "group", # Node group, chr
             # Nodesize = "label" # chr !!!
             # linkColour = "value"
             )



```


### Creating a Sankey Diagram. 


```{r sankeyNetwork}
UCB_graph <-
  UCBAdmissions %>% 
  as.data.frame() %>% 
  select(Gender, Admit, Dept, Freq) %>% 
  as_tbl_graph()
UCB_graph

UCB_nodes <- UCB_graph %>% activate(nodes) %>% as_tibble()
UCB_nodes

UCB_links <- UCB_graph %>% 
  activate(edges) %>% 
  as_tibble() %>%
  dplyr::mutate(from = from - 1, to = to - 1)
UCB_links

sankeyNetwork(
  Links = UCB_links,
  Nodes = UCB_nodes,
  Source = "from",
  Target = "to",
  Value = "Freq",
  LinkGroup = "Dept",
  fontSize = 20,
  fontFamily = "Arial"
)
```

`simpleNetwork`, `forceNetwork` and `sankeyNetwork` use a similar `node` and `link` data structure, organized as two **data frames** (not tibbles)

` chordNetwork` uses an *association matrix* type of matrix or a data frame organized in the same way, where entry (n,m) represents the strength of the link from group n to group m. Matrix needs to be **square** !! "Column names" and "Row names" need to be the same if the data is a `data.frame`.

```{r chordNetwork}
data <- matrix(rpois(n = 16, lambda = 50), nrow = 4, ncol = 4)
chordNetwork(data,
             labels = c("A", "B", "C", "D"))

```




## Using threejs

1. <https://bwlewis.github.io/rthreejs/>
2. <https://bwlewis.github.io/rthreejs/crosstalk.html>

### `graphjs` usage

```{r}
data("LeMis") # igraph object
LeMis
V(LeMis)$label
V(LeMis)$color
graphjs(LeMis,
    layout=list( # animates between a list of layouts
      # layouts need to be 3D layouts
      # Or each can be a 3 column matrix with n(rows) = n(vertices)
    #layout_randomly(LeMis, dim=3),
    #layout_on_sphere(LeMis),
    layout_with_drl(LeMis, dim=3)  # note! somewhat slow...
    #layout_with_fr(LeMis, dim=3, niter=30)
    ),
  main=list("random layout", "sphere layout", "drl layout", "fr layout"),
  fpl=300)

```


## `globejs` usage
Plot points, arcs and images on a globe in 3D using Three.js. The globe can be rotated and and zoomed. Great Circles and historical routes are a good idea for this perhaps.

Refer to this page for more ideas <http://bwlewis.github.io/rthreejs/globejs.html>

```{r}
# Random Lats and Longs
lat <- rpois(10, 60) + rnorm(10, 80)
long <- rpois(10, 60) + rnorm(10, 10)

# Random "Spike" heights for each location. Population? Tourists? GDP?
value <- rpois(10, lambda = 80)
 
globejs(lat = lat, long = long)

globejs(
  lat = lat,
  long = long,
  
  # random heights of the Spikes (!!) at lat-long combo
  value = value,
  color = "red",
  # Zoom factor, default is 35
  fov = 50
)

globejs(
  lat = lat,
  long = long,
  value = value,
  color = "red",
  pointsize = 4, # width of the columns
  # Zoom position
  fov = 35,
  # initial position of the globe
  rotationlat = 0.6, #  in RADIANS !!! Good Heavens!!
  rotationlong = 0.2 #  in RADIANS !!! Good Heavens!!
)

globejs(
  lat = lat,
  long = long,
  value = value,
  color = "red",
  pointsize = 4,
  fov = 35,
  rotationlat = 0.6,
  rotationlong = 0.2,
  lightcolor = "#aaeeff",
  emissive = "#0000ee",
  bodycolor = "#ffffff",
  bg = "grey"
)

```


```{r}
# Plotting arcs on the globe
# Requires 4 column data frame for start and end lat and long

arcs <- data.frame(start_lat = runif(10, min = -20, max = 20), 
                   start_lon = runif(10, min = -20, max = 20), 
                   end_lat = runif(10, min = -20, max = 20),
                   end_lon = runif(10, min = -20, max = 20) + 60)
arcs

globejs(arcs = arcs,
        arcsColor = "gold",
        arcsLwd = 4,
        arcsHeight = 0.6,
        rotationlat = 0,
        rotationlong = -2.2)
```

`threejs` contains a dataset called `flights` obtained from Callum Prentice's FlightStream page:
< http://callumprentice.github.io/apps/flight_stream/index.html> 

```{r}
data("flights",package = "threejs")
flights %>% head()

# Example Code
# dest   <- factor(sprintf("%.2f:%.2f",flights[,3], flights[,4]))
# dest
# freq <- sort(table(dest), decreasing=TRUE)
# freq
frequent_destinations <-
  flights %>% group_by(dest_lat, dest_long) %>% 
  count(sort = TRUE) %>% 
  ungroup() %>% 
  slice_max(n = 10, order_by = n)
frequent_destinations

# idx <- dest %in% frequent_destinations
# frequent_flights <- flights[idx, ]
# ll <- unique(frequent_flights[,3:4])
frequent_flights <- flights %>% 
  semi_join(frequent_destinations,by = c("dest_lat" = "dest_lat", "dest_long" = "dest_long")) %>% unique() 

frequent_flights %>% 
  kableExtra::kbl() %>%
  kableExtra::kable_paper(full_width = TRUE) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "condensed", "responsive")) %>%
  kableExtra::scroll_box(width = "700px", height = "500px")


# globejs(lat=ll[, 1], long=ll[, 2], arcs=frequent_flights,
#         bodycolor="#aaaaff", arcsHeight=0.3, arcsLwd=2,
#         arcsColor="#ffff00", arcsOpacity=0.15,
#         atmosphere=TRUE, color="#00aaff", pointsize=0.5)

# Get a nice image of the globe from NASA
earth <- "http://eoimages.gsfc.nasa.gov/images/imagerecords/73000/73909/world.topo.bathy.200412.3x5400x2700.jpg"

globejs(
  lat = frequent_flights$dest_lat,
  long = frequent_flights$dest_long,
  arcs = frequent_flights,
  #value = frequent_destinations$n,
  color = "red",
  #bodycolor = "#aaaaff",
  arcsHeight = 0.3,
  arcsLwd = 2,
  arcsColor = "#ffff00",
  arcsOpacity = 0.15,
  atmosphere = FALSE,
  #color="#00aaff",
  pointsize = 2,
  rotationlat = 0.6,
  rotationlong = 0.2,
  # lightcolor = "#aaeeff",
  # emissive = "#0000ee",
  # bodycolor = "#ffffff",
  bg = "grey"
        )
```

## Using `scatterplot3js` and friends

3D scatter plots with points and lines can be achieved using `scatterplot3js`, `points3D`, and `lines3D`. 

```{r}
library(palmerpenguins)
penguins <- penguins %>% drop_na()
scatterplot3js(x = penguins$bill_length_mm, 
               y = penguins$flipper_length_mm, 
               z = penguins$body_mass_g)

```



```{r}
library(palmerpenguins)
penguins <- penguins %>% drop_na()
scatterplot3js(x = penguins$bill_length_mm, 
               y = penguins$flipper_length_mm, 
               z = penguins$body_mass_g,
               cex.symbols = 0.2) # Smaller Points

```



```{r}
library(palmerpenguins)
penguins <- penguins %>% drop_na()

scatterplot3js(x = penguins$bill_length_mm, 
               y = penguins$flipper_length_mm, 
               z = penguins$body_mass_g,
               cex.symbols = 0.2) %>%  # Smaller Points
lines3d(from = penguins$bill_length_mm, 
        to = penguins$flipper_length_mm, 
        color = "blue")

```



# Using `leaflet`

`leaflet` allows one to create interactive maps with very simple and intuitive code. `leaflet` also includes access to many "base maps" that can change the look and feel of the map itself, which can be very useful ( e.g. night mode maps). Information on how to use `leaflet` is available at <http://rstudio.github.io/leaflet/>

```{r}
leaflet() %>% 
  addTiles() %>% 
  setView(lng = 77, lat = 18, zoom = 4)
```

