---
title: "\U0001F578 Networks"
subtitle: "How Things Connect and Work Together"
author: "Arvind Venkatadri"
date: 12/July/2022
lastmod: "`r Sys.Date()`"
weight : 100
summary: "Networks and Connections and what happens over them"
tags:
- Nodes and Edges
- Network Diagrams
- Tree Diagrams
- Arc Diagrams
image:
  placement: 2
  caption: Photo by <a href="https://unsplash.com/@designedbyflores?utm_source=unsplash&utm_medium=referral&utm_content=creditCopyText">Omar Flores</a> on <a href="https://unsplash.com/photos/MOO6k3RaiwE?utm_source=unsplash&utm_medium=referral&utm_content=creditCopyText">Unsplash</a>
  
  focal_point: ""
  preview_only: true
projects: []
links:
- icon: flask
  icon_pack: fas
  name: lab
  url: /labs/R-for-Artists/07-networks/07-Grammar-of-Networks.html
#   
# - icon: magic
#   icon_pack: fas
#   name: Introduction Slides
#   url: https://r-for-artists.netlify.app/slides/06-Spatial-Data-in-R.html
---


## Introduction

What is a network?

Network graphs show relationships between entities: what sort they are,
how strong they are, and even of they change over time.

We will examine data structures pertaining both to the *entities* and
the *relationships* between them and look at the data *object* that can
combine these aspects together. Then we will see how these are plotted,
what the *structure* of the plot looks like. There are also *metrics*
that we can calculate for the network, based on its structure. We will
of course examine geometric metaphors that can represent various classes
of entities and their relationships.

Network graphs can be rendered both as static and interactive and we
will examine R packages that render both kinds of plots.

There is a another kind of structure: one that combines *spatial* and
*network* data in one. We will defer that for a future module !


`r blogdown::shortcode("youtube", "DrhhdekP1ZQ")`


## Packages that we will need


```{r setup, include=TRUE}
#| label: setup
#| include: true
#| message: false
#| warning: false
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

library(tidygraph) # Creating Graph objects and analyzing them
library(ggraph) # Plotting graphs with various layout options
library(graphlayouts) # More layout options
library(igraphdata) # Sample graph datasets

library(sand) # Statistical Analysis of Graph Data


```



## What kind Network graphs will we make?

Here is a network map of the characters in Victor Hugo's
[`Les Miserables`](https://www.gutenberg.org/files/135/135-h/135-h.htm):

<iframe width="100%" height="676" frameborder="0" src="https://observablehq.com/embed/@d3/force-directed-graph-canvas?cells=chart">

</iframe>

And this:

```{r}
#| echo: false
#| warning: false
data(karate)
set.seed(123)
l <- layout.graphopt(karate)
karate %>% 
  as_tbl_graph() %>% # to access nodes and edges in tidygraph
  
  # parameters to decorate the nodes
  activate(nodes) %>% 
  mutate(node_shape = if_else(V(karate)$label %in% c("H", "A"), 1, 2),
         node_size = 4*sqrt(igraph::degree(karate)),
         Faction = as_factor(Faction),
         node_shape = as_factor(node_shape)
         ) %>% 
  
  # parameters to decorate the edges
  activate(edges) %>%
  mutate(edge_color =
           case_when(
             edge_is_between(from = V(karate)$Faction == 1,
                             to = V(karate)$Faction == 2,
                             ignore_dir = TRUE) ~ 1L,
             edge_is_between(from = V(karate)$Faction == 1,
                             to = V(karate)$Faction == 1,
                             ignore_dir = TRUE) ~ 2L,
             edge_is_between(from = V(karate)$Faction == 2,
                             to = V(karate)$Faction == 2,
                             ignore_dir = TRUE) ~ 3L,
             TRUE ~ 4L),
         edge_color = as_factor(edge_color)) %>%
  
  # now plot it
  ggraph(., layout = l) +
  geom_edge_link0(aes(width = weight, 
                      color = edge_color)) + 
  geom_node_point(aes(size = node_size,
                      color = Faction,
                      shape = node_shape
                      )) +
  geom_node_text(aes(label = 
                       sub("Actor","",V(karate)$name)),repel = TRUE) + 
  scale_size(range = c(2, 10)) +
  scale_colour_manual(values = c("red", "dodgerblue")) +
  scale_shape_manual(values = c(15, 19)) +
  scale_edge_width(range = c(0.1, 4)) +
  scale_edge_colour_manual(values = c("yellow", "pink", "lightblue", "grey")) +
  theme_graph() +   
  labs(title = "Zachary's Karate Club Membership Network") +
  theme(legend.position = "none", title = element_text(size = 6))


```

## Conclusion

We see that Networks reveal *Structure* of relationships between entities, their types and strengths, and also reveal behavioural aspects as to which entities or relationships are important which are peripheral to the overall network.

## References

1.  [Hadley Wickham](http://hadley.nz/), [Danielle
    Navarro](https://djnavarro.net/), and [Thomas Lin
    Pedersen](https://www.data-imaginist.com/), *ggplot2: Elegant Graphics for Data Analysis*.<https://ggplot2-book.org/networks.html>

2.  Omar Lizardo and Isaac Jilbert, *Social Networks: An Introduction*.
    <https://bookdown.org/omarlizardo/_main/>

3.  Mark Hoffman, *Methods for Network Analysis.*
    <https://bookdown.org/markhoff/social_network_analysis/>

4.  Eric D. Kolaczyk and Gábor Csárdi. *Statistical Analysis of Network Data with R, 2nd
    Edition*.<https://github.com/kolaczyk/sand>
    
    
