---
title: "Slides 08 - The Grammar of Diagrams"
subtitle: ""  
author: "Arvind Venkatadri"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    css: ["default", "css/ohsu.css", "css/ohsu-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: atelier-lakeside-light
      highlightLines: true
      slideNumberFormat: "%current%"
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, comment = NA, dpi = 300,fig.align = "center", cache = FALSE, fig.path= "08-figs/")
library(xaringanthemer)
library(xaringanExtra)
library(tidyverse)
library(DiagrammeR)
library(nomnoml)
```


```{r xaringanExtra, include=FALSE}
use_webcam()
use_fit_screen()
use_panelset()
```


# Let's see how to use DiagrammeR

- `DiagrammeR` uses the processing function called `grViz()`
- What you pass into grViz() is a valid graph specification in the **DOT language**. 
- The DOT graph description is delivered to grViz() in the form of a string, enclosed in quotes.

---

# Let's see how to use DiagrammeR

Diagrams can be **undirected**...

.pull-left[
```{r basics1,  message=FALSE, eval=FALSE, warning=FALSE}

DiagrammeR::grViz(
  "graph {A -- B}" #<<
  )

```
]

.pull-right[

```{r ref.label='basics1', echo=FALSE,message=FALSE,warning=FALSE, out.width= "50%", out.height="50%", fig.align='center'}

```
]

---

# Let's see how to use DiagrammeR

Diagrams can be undirected...or **directed**

.pull-left[
```{r basics2,  message=FALSE, eval=FALSE, warning=FALSE}

DiagrammeR::grViz(
  "digraph {A -> B}" #<<
  )

```
]

.pull-right[

```{r ref.label='basics2', echo=FALSE,message=FALSE,warning=FALSE, out.width= "50%", out.height="50%", fig.align='center'}

```
]

---

# Nodes and Labels

.pull-left[

```{r big,message=FALSE, eval=FALSE, warning=FALSE }
grViz("
digraph boxes_and_circles {

  # a 'graph' statement
  graph [overlap = true, fontsize = 10]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica]
  A; B; C; D; E; F

  node [shape = circle,
        fixedsize = true,
        width = 0.9] // sets as circles
  1; 2; 3; 4; 5; 6; 7; 8

  # several 'edge' statements
  A->1 B->2 B->3 B->4 C->A
  1->D E->A 2->4 1->5 1->F
  E->6 4->6 5->7 6->7 3->8
}
")

```
]


.pull-right[

```{r ref.label='big', echo=FALSE,message=FALSE,warning=FALSE, out.width= "50%", out.height="50%", fig.align='center'}

```

]
