---
title: "Chap03Testing_Exer"
author: "Chihara-Hesterberg"
date: "December 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width="50%")
library(dplyr)
library(ggplot2)
```

##Chapter 3 Introduction to Hypothesis Testing: Permutation Tests
##Exercises

###Exercise 7
LaGuardia Airport (LGA) is one of three major airports that serves the
New York City metropolitan area. In 2008, over 23 million passengers and
over 375 000 planes flew in or out of LGA. United Airlines and American
Airlines are two major airlines that schedule services at LGA. The data set
FlightDelays contains information on all 4029 departures of these two
airlines from LGA during May and June 2009.

```{r}
FlightDelays <- read.csv("http://sites.google.com/site/chiharahesterberg/data2/FlightDelays.csv")
```

N<-10^4-1
UA.Delay <- FlightDelays %>% filter(Carrier == "UA") %>% pull(Delay)
AA.Delay <- FlightDelays %>% filter(Carrier == "AA") %>% pull(Delay)

observedSumUA <- sum(UA.Delay)
observedmeanUA <- mean(UA.Delay)
observedmeanDiff <- mean(UA.Delay) - mean(AA.Delay)
m <-length(UA.Delay)  #number of UA observations

sumUA<-numeric(N)
meanUA<-numeric(N)
meanDiff<-numeric(N)

set.seed(0)
for (i in 1:N)
{
  index <- sample(4029,  m, replace = FALSE)
  sumUA[i] <- sum(FlightDelays$Delay[index])
  meanUA[i] <- mean(FlightDelays$Delay[index])
  meanDiff[i] <- mean(FlightDelays$Delay[index]) - mean(FlightDelays$Delay[-index])

}

(sum(sumUA >= observedSumUA) + 1)/(N + 1)  #P-value

(sum(meanUA >= observedmeanUA) + 1)/(N + 1)  #P-value

(sum(meanDiff >= observedmeanDiff) + 1)/(N + 1)  #P-value

```
