---
title: "06-Making-Maps-with-tmap"
subtitle: "Using the tmap package" 
author: "Arvind Venkatadri"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: 
      - default
      - css/lexis.css
      - css/lexis-fonts.css
    lib_dir: libs
    nature:
      highlightStyle: atelier-lakeside-light
      highlightLines: true
      slideNumberFormat: "%current%"
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE,warning=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, comment = NA, dpi = 300,fig.align = "center", cache = FALSE)
library(xaringanthemer)
library(xaringanExtra)
library(tidyverse)
library(sf)
library(tmap)
library(spData)
#install.packages("spDataLarge", repos = "https://nowosad.github.io/drat/", type = "source")
library(spDataLarge)
library(rnaturalearth)
library(rnaturalearthdata)
#
library(patchwork)
library(emoji)
library(emojifont)
library(anicon)
library(here)
tmap_mode("plot")
```


```{r xaringanExtra, include=FALSE}
use_webcam()
use_fit_screen()
use_panelset()
use_broadcast()
use_extra_styles(hover_code_line = TRUE,mute_unhighlighted_code = FALSE)
```


class: center, middle

# Map Making with `tmap`
### Adapted from SIGR2021 Conference (*Sciences de l'information géographique reproductibles*)
#### Saint-Pierre-d'Oléron (France)
#### 27/June/2021 to 3/July/2021
#### "GIS and mapping": "A workshop on GIS and mapping with R: part 2"
#### Jakub Nowosad & Robin Lovelace

---

# Mapping

- We have seen a bit of `ggplot + geom_sf + coord_sf()`
- This is good for static plots
- `tmap` can give us both static and interactive plots
- And some fancy map templates too!!

---


# Mapping

Let's look at London Boroughs and places where we can hire bicycles.

- The `lnd` dataset in the package `spData` will give us a London Map.

- The `cycle_hire_osm` dataset, also in the package `spData` will give us Cycle Hire Locations in London.
- What `GEOMETRY` column types to you expect in both of these datasets?

--
  - MULTIPOLYGON?
  
--
  - ~~POLYGON?~~
  
--
  - ~~MULTILINE?~~
  
--
  - ~~LINE~~
  
--
  - ~~MULTIPOINT?~~
  
--
  - POINT?
  

---

# London

```{r}

data("lnd", package = "spData") # Boroughs of London
str(lnd)

```

---

# Cycle Hire Locations in London

```{r}

data("cycle_hire_osm", package = "spData") # Cycle Hire locations
str(cycle_hire_osm)

```

---
# Step-by-Step `tmap` Code

.pull-left[
```{r tm10, warning=FALSE, fig.show='hide', message=FALSE}

tm_shape(lnd) + #<<
  tm_borders(col = "black", lwd = 1) + #<<
  tm_fill("NAME", legend.show = FALSE) #<<

```

]

.pull-right[
```{r, echo=FALSE}

knitr::include_graphics(
  knitr::fig_chunk("tm10", "png")
)

```
]

---
# Step-by-Step `tmap` Code

.pull-left[
```{r tm20, warning=FALSE, fig.show='hide', message=FALSE}

tm_shape(lnd) + 
  tm_borders(col = "black", lwd = 1) +
  tm_fill("NAME", legend.show = FALSE,
          palette = "Accent") #<<

```

- The default color can be changed using the `palette` argument 
- run the `tmaptools::palette_explorer()` function in your `CONSOLE`to see possible palette names.
- Refer to <https://geocompr.github.io/post/2019/tmap-color-scales/>
]

.pull-right[
```{r, echo=FALSE}

knitr::include_graphics(
  knitr::fig_chunk("tm20", "png")
)

```
]

---

# `tmap` Layers are in *groups*


.pull-left[

Groups start with `tm_shape(data = ...)`   
Groups are "connected" with a $+$ sign as usual...

```{r tm30, warning=FALSE, fig.show='hide', message=FALSE}

# Group 1 #<<
tm_shape(lnd) +
  tm_borders(col = "black",lwd = 1) + 
  tm_fill("NAME", legend.show = FALSE) +

# Group 2 = Layer 2 #<<
  tm_shape(cycle_hire_osm) + #<<
  tm_symbols(size = 0.2, col = "red") #<<

```
]

.pull-right[
```{r, echo=FALSE}

knitr::include_graphics(
  knitr::fig_chunk("tm30", "png")
)

```
]

---
# Add graticules and legends

.pull-left[
```{r tm40, warning=FALSE, fig.show='hide', message=FALSE}

# Group 1 #<<
tm_shape(lnd) +
  tm_borders(col = "black",lwd = 1) + 
  tm_fill("NAME", legend.show = FALSE) +
# Group 2 = Layer 2 #<<
  tm_shape(cycle_hire_osm) + 
  tm_symbols(size = 0.2, col = "red") +
  
  tm_graticules() + #<<
  tm_add_legend(type = "symbol", #<<
                col = "red", #<<
                title = "Hire Cycles Here") #<<

```

- Still 2 groups only !!
]

.pull-right[
```{r, echo=FALSE}

knitr::include_graphics(
  knitr::fig_chunk("tm40", "png")
)

```
]

---
# Add Scale and Compass to the Map

.pull-left[

```{r tm50, warning=FALSE, fig.show='hide', message=FALSE}

# Group 1
tm_shape(lnd) +
  tm_borders(col = "black",lwd = 1) + 
  tm_fill("NAME", legend.show = FALSE) +
# Group 2 = Layer 2
  tm_shape(cycle_hire_osm) + 
  tm_symbols(size = 0.2, col = "red") +
  
  tm_graticules() +
  tm_add_legend(type = "symbol", 
                col = "red",
                title = "Hire Cycles Here") +

  tm_scale_bar(position=c("left", "bottom"),#<<
               text.size = 1) + #<<
  tm_compass(position = c("right", "top"), #<<
             type = "rose", #<<
             size = 2)  #<<

```
]

.pull-right[
```{r, echo=FALSE}

knitr::include_graphics(
  knitr::fig_chunk("tm50", "png")
)

```
]

---
# Add Credits and Layout Options

.pull-left[

```{r tm60, warning=FALSE, fig.show='hide', message=FALSE}

# Group 1
tm_shape(lnd) +
  tm_borders(col = "black", lwd = 1) + 
  tm_fill("NAME", legend.show = FALSE) +
# Group 2 = Layer 2
  tm_shape(cycle_hire_osm) + 
  tm_symbols(size = 0.2, col = "red") +
  tm_graticules() +
  tm_add_legend(type = "symbol", col = "red", title = "Hire Cycles Here") +
  tm_scale_bar(position=c("left", "bottom"),text.size = 1) + 
  tm_compass(position = c("right", "top"), type = "rose", size = 2) +
  tm_credits(text = "Arvind V, 2021", #<<
             position = c("left", "bottom")) + tm_layout(main.title = "London Bike Share", #<<
bg.color = "lightblue", #<<
inner.margins = c(0, 0, 0, 0)) -> mymap #<<

```
]

.pull-right[
```{r, echo=FALSE}

knitr::include_graphics(
  knitr::fig_chunk("tm60", "png")
)

```
]



---

# Saving

```{r tm80, eval=FALSE}

tmap_save(my_map, filename = "my_map.png", width = 300, height = 800, dpi = 300)

```

---

# And..Interactive?
.leftcol30[
- Use `tmap_mode("plot")` for static plots (default)
- Use `tmap_mode("view")` for interactive plots!
- Click on the layer button to choose other background base maps!!
]

.rightcol70[

```{r tm70, warning=FALSE, message=FALSE, out.height="400px", out.width="750px"}
tmap_mode("view")
mymap

```
]


---

# Thanks!

Slides created by Arvind Venkatadri via the R packages:

[**xaringan**](https://github.com/yihui/xaringan)<br>

