---
title: "\U0001F355 Parts of a Whole"
subtitle: "Parts of a Whole"
author: "Arvind Venkatadri"
date: "`r Sys.Date()`"
weight: 50
summary: "Slices, Portions, Counts, and Aggregates of Data"
tags:
  - Pie Charts/Fan Charts/Donut Charts
  - Mosaic Charts
  - Venn Diagrams
  - Grouped/Stacked/Circular Bar Charts
  - Tree Maps/Dendrograms
  - Parliament/Waffle Charts
  - Circular Packing
  - Pyramid Plots

image:
  caption: Photo by rawpixel on Unsplash
  focal_point: Smart
external_link: ""
links:
- icon: circle
  icon_pack: fas
  name: Orange Tutorial
  url: "labs/Data-Analytics/parts.ows"
- icon: sun
  icon_pack: fas
  name: Radiant
  url: "labs/Data-Analytics/parts.rda"
- icon: r-project
  icon_pack: fab
  name: R Tutorial
  url: "labs/Data-Analytics/parts.html"
- icon: database
  icon_pack: fas
  name: Sample Datasets
  url: "labs/Data-Analytics/data/qdd-data.zip"
slides: ""
url_code: ""
url_pdf: ""
url_slides: ""
url_video: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(blogdown)
library(plotrix) # Fan, Pyramid Chart
library(ggparliament) # Parliament Chart
library(waffle) # What it says!

```

## Introduction

So let us start with "eating humble pie": ~creating~ discussing a Pie chart first.

A `pie chart` is a circle divided into sectors that each represent a proportion of the whole. It is often used to show percentage, where the sum of the sectors equals 100%. The problem is that humans are pretty bad at reading angles.  This ubiquitous chart is much vilified in the industry and  `bar charts` that we have seen earlier, are viewed as better options. However do read this spirited defense of pie charts here. <https://speakingppt.com/why-tufte-is-flat-out-wrong-about-pie-charts/>

On the other hand, pie charts are ubiquitous in business circles, so there is an attractive alternative, called a `fan chart` which we will explore here. 

We will try to depict here the interesting ones such as the `dendrogram`, the `parliament plot`, the `waffle plot`, and the `circle packing` chart. 

### Fan Plot

Base R has a simple `pie` command that does the job.

```{r}
pie.sales <- c(0.12, 0.3, 0.26, 0.16, 0.04, 0.12)
labels <- c("Blueberry", "Cherry",
    "Apple", "Boston Cream", "Other", "Vanilla Cream")
pie(x = pie.sales, labels = labels,col =  grDevices::hcl.colors(palette= "Plasma", n = 6)) # default colours

```
The `fan plot` displays numerical values as arcs of **overlapping sectors**. This allows for more effective comparison: 

```{r,echo=TRUE}
plotrix::fan.plot(x = pie.sales, labels = labels,
                  col = grDevices::hcl.colors(palette= "Plasma", n = 6),
                  shrink = 0.03, # How much to shrink each successive sector
                  label.radius = 1.15,
                  main = "Fan Plot of Ice Cream Flavours",
                  ticks = 360,
                  max.span = pi)


```

## Parliament and Waffle Charts

```{r}
parts <- c(80, 30, 20, 10)
waffle(parts, rows=8)

```

