---
title: "\U0001F576 Happy Data are all Alike"
subtitle: "Distributions"
author: "Arvind Venkatadri"
date: "`r Sys.Date()`"
weight: 20
summary: "Single Quant Variable Graphs and their Siblings"
tags:
- Qual Variables and Quant Variables
- Bar and Column Charts
- Histograms and Density Plots

image:
  caption: Photo by rawpixel on Unsplash
  focal_point: Smart
external_link: ""
links:
# - icon: linkedin
#   icon_pack: fab
#   name: 
#   url: https://linkedin.com/arvindvenkatadri
# - icon: twitter
#   icon_pack: fab
#   name: 
#   url: https://twitter.com/arvind_v
slides: ""
url_code: ""
url_pdf: ""
url_slides: ""
url_video: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(blogdown)
library(tidyverse)
library(downloadthis)

```

```{css, echo=FALSE}
/* from https://ianlunn.github.io/Hover/ */
.hvr-sweep-to-left {
  display: inline-block;
  vertical-align: middle;
  -webkit-transform: perspective(1px) translateZ(0);
  transform: perspective(1px) translateZ(0);
  box-shadow: 0 0 1px rgba(0, 0, 0, 0);
  position: relative;
  -webkit-transition-property: color;
  transition-property: color;
  -webkit-transition-duration: 0.3s;
  transition-duration: 0.3s;
}
.hvr-sweep-to-left:before {
  content: "";
  position: absolute;
  z-index: -1;
  top: 0;
  left: 0;
  right: 0;
  bottom: 0;
  background: #2098D1;
  -webkit-transform: scaleX(0);
  transform: scaleX(0);
  -webkit-transform-origin: 100% 50%;
  transform-origin: 100% 50%;
  -webkit-transition-property: transform;
  transition-property: transform;
  -webkit-transition-duration: 0.3s;
  transition-duration: 0.3s;
  -webkit-transition-timing-function: ease-out;
  transition-timing-function: ease-out;
}
.hvr-sweep-to-left:hover, .hvr-sweep-to-left:focus, .hvr-sweep-to-left:active {
  color: white;
}
.hvr-sweep-to-left:hover:before, .hvr-sweep-to-left:focus:before, .hvr-sweep-to-left:active:before {
  -webkit-transform: scaleX(1);
  transform: scaleX(1);
}

```

## What graphs will we see today?

Some of the very basic and commonly used plots for data are:
- Bar and Column Charts
- Histograms and Frequency Distributions
- Scatter Plots (if there is more than one quant variable)
and
- 2D Hexbins Plots and 2D Frequency Distributions (horrors?)

## An Example: Bar and Column Charts

Here is a dataset from Jeremy Singer-Vine's blog, [Data Is Plural](https://www.data-is-plural.com/). This is a list of all books banned in schools across the US. 

```{r, echo=FALSE}
banned <- readxl::read_xlsx(path = "./banned.xlsx",
                            sheet = "Sorted by Author & Title") %>%
  janitor::clean_names()

banned
banned %>% download_this(output_name = "banned", output_extension = ".csv", button_label = "Banned Books data", button_type = "default", icon = "fa fa-save")

``` 


```{r}
banned_by_state <- banned %>% 
  group_by(state) %>% 
  summarise(total = n())
banned_by_state

banned %>% 
  group_by(state, type_of_ban) %>% 
  summarise(count = n()) %>%
  slice_max(order_by = count,n = 10) %>%
 #  pivot_wider(.,id_cols = State,
 #              names_from = `Type of Ban`,
 #              values_from = count) %>% janitor::clean_names() %>% 
 #  replace_na(list(banned_from_libraries_and_classrooms = 0,
 #                  banned_from_libraries = 0,
 #                  banned_pending_investigation = 0,
 #                  banned_from_classrooms = 0)) %>% 
 # mutate(total = sum(across(where(is.integer)))) %>%
ggplot(aes(x = reorder(state, count), y = count, fill = type_of_ban)) + geom_col() + labs(title = "Banned Books by Type across the US") + xlab("State") + coord_flip() + theme_classic()

```


## An Example: Histograms and Frequency Distributions

TBD: Example using Flourish

### How does this Work?

Histograms are best to show the distribution of raw *quantitative data*, by displaying the number of values that fall within defined ranges, often called *buckets* or *bins*. 

Although histograms may look similar to column charts, the two are different. First, histograms show *continuous data*, and usually you can adjust the bucket ranges to explore frequency patterns. For example, you can shift histogram buckets from 0-1, 1-2, 2-3, etc. to 0-2, 2-4, etc. 
By contrast, column charts show *categorical data*, such as the number of apples, bananas, carrots, etc. Second, histograms do not usually show spaces between buckets because these are *continuous values*, while column charts show spaces to separate each *category*.


### How could you explore?

TBD. Add `hexbin` plots here

### What is the Story here?

TBD




## An Example: Frequency Density

### How does this work?

Let us listen to the late great Hans Rosling from the [Gapminder Project](https://www.gapminder.org), which aims at telling stories of the world with data, to remove systemic biases about poverty, income and gender related issues.


<iframe src="https://player.vimeo.com/video/127511255?h=a225cf98cc" width="640" height="360" frameborder="0" allow="autoplay; fullscreen; picture-in-picture" allowfullscreen></iframe>
<p><a href="https://vimeo.com/127511255">How many are rich and how many are poor?</a> from <a href="https://vimeo.com/gapminder">Gapminder</a> on <a href="https://vimeo.com">Vimeo</a>.</p>

<p align="center"><iframe src="//www.gapminder.org/tools/?embedded=true#$model$markers$mountain$encoding$frame$speed:187;;;;;&chart-type=mountain&url=v1" style="width: 100%; height: 500px; margin: 0 0 0 0; border: 1px solid grey;"allowfullscreen></iframe></p>

### How could you explore?
TBD. Add 2D contour plots and link up to `hexbin plots`.

### What is the Story here?


## Your Turn

```{r echo=FALSE}
airbnb <- read.table("https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/1_OneNum.csv", header=TRUE)

apartments <- read.table("https://raw.githubusercontent.com/holtzy/data_to_viz/master/Example_dataset/2_TwoNum.csv", header=T, sep=",") %>% select(GrLivArea, SalePrice)

```

1. Rbnb Price Data on the French Riviera: 
```{r, echo = FALSE}
airbnb %>% download_this(output_name = "airbnb",output_extension = ".csv", button_label = "Rbnb data", button_type = "default", icon = "fa fa-save")
```  


1. Apartment price vs ground living area: 
```{r, echo = FALSE}
apartments %>% download_this(output_name = "apartments", output_extension = ".csv", button_label = "Apartment Data", button_type = "default", icon = "fa fa-save",   class = "hvr-sweep-to-left")
```
(Try a Scatter Plot too, since we have two Quant variables)  

1. Rbnb Price Data on the French Riviera: 
```{r, echo=FALSE}
airbnb %>% download_this(output_name = "airbnb",output_extension = ".csv", button_label = "Rbnb data",button_type = "default", icon = "fa fa-save", class = "hvr-sweep-to-left")
```  


1. India 
1. Old Faithful Data
1. Income data
3. Diamonds Data from R
4. calmcode.io dataset


## Fun Stuff

1. See the scrolly animation for a histogram at this website: **Exploring Histograms, an essay by Aran Lunzer and Amelia McNamara**
<https://tinlizzie.org/histograms/?s=09>

