---
title: "\U0001F579 Eating Mangoes with Hamlet"
author: "Arvind Venkatadri"
date: "`r Sys.Date()`"
weight: 75
external_link: ""
image:
  caption: Tom und Nicki Löschner on Pixabay
  focal_point: Smart
links:
slides: 
summary: 
tags:
- Hamlet
- Hypothesis
- Design of Experiments
- PPDAC Workflow
- Sampling
- Randomization
url_code: ""
url_pdf: ""
url_slides: ""
url_video: ""
editor_options: 
  markdown: 
    wrap: 72
---

```{r include=FALSE}
library(blogdown)
library(tidyverse)
library(mosaic)
library(patchwork)
library(ggtext)
library(ggformula)
```

## Hamlet is also, yeah, I mean,... like us! Random!!

`r blogdown::shortcode("youtube", "EmjWIMMtjDc")` 
<br>

Let us read the famous nunnery scene from <u>[**Hamlet: To be or Not to
be**...](https://www.nosweatshakespeare.com/quotes/hamlet-to-be-or-not-to-be/)</u>

**Hamlet’s ‘To Be Or Not To Be’ Speech, Act 3 Scene 1**

> To be, or not to be: that is the question:\
> Whether ’tis nobler in the mind to suffer\
> The slings and arrows of outrageous fortune,\
> Or to take arms against a sea of troubles,\
> And by opposing end them? To die: to sleep;\
> No more; and by a sleep to say we end\
> The heart-ache and the thousand natural shocks\
> That flesh is heir to, ’tis a consummation\
> Devoutly to be wish’d. To die, to sleep;\
> To sleep: perchance to dream: ay, there’s the rub;\
> For in that sleep of death what dreams may come\
> When we have shuffled off this mortal coil,\
> Must give us pause: there’s the respect\
> That makes calamity of so long life;\
> For who would bear the whips and scorns of time,\
> The oppressor’s wrong, the proud man’s contumely,\
> The pangs of despised love, the law’s delay,\
> The insolence of office and the spurns\
> That patient merit of the unworthy takes,\
> When he himself might his quietus make\
> With a bare bodkin? who would fardels bear,\
> To grunt and sweat under a weary life,\
> But that the dread of something after death,\
> The undiscover’d country from whose bourn\
> No traveller returns, puzzles the will\
> And makes us rather bear those ills we have\
> Than fly to others that we know not of?\
> Thus conscience does make cowards of us all;\
> And thus the native hue of resolution\
> Is sicklied o’er with the pale cast of thought,\
> And enterprises of great pith and moment\
> With this regard their currents turn awry,\
> And lose the name of action.–Soft you now!\
> The fair Ophelia! Nymph, in thy orisons\
> Be all my sins remember’d.\

What was Hamlet's problem? How did he solve that problem?

## Suppose we were Hamlet....

What would we want to know?

-   Who is our friend?
-   What is the chance that we will survive?
-   Who killed Dad?

Did Hamlet have an idea, any suspicion, that he might have wanted to
prove? Yes, he did, and that is why he *feigned madness*, so people
would open up and tell him things, in an unguarded way. So there was
<u>[**craft in his
madness**](http://www.shakespeare-online.com/plays/hamlet/antichamlet.html)!!</u>\

<u>[**Hakuna Matata, Dude!!**](https://youtu.be/kaOIxll4LCA)</u>

### A Hypothesis worthy of Hamlet

If we wanted to know the truth of something, some guess, a hunch, or a
**Hypothesis** that we had, then we would need to conduct an
"Experiment", rather like Hamlet did, without quite going or **feigning
madness** of course. But there is a **Madness to/in the Method**, as we
will see...

!["Design of Experiments - A Workflow: A Madness in the
Method"](https://www.statsref.com/HTML/ppdac.png)

From
<u>[**StatsRef**](https://www.statsref.com/HTML/?statistics__statistical_analys.html)</u>
:

> The PPDAC summary table suggests a relatively linear flow from problem
> definition/Hypothesis through to conclusions --- this is typically not
> the case. It is often better to see the process as cyclical, with a
> series of feedback loops. A summary of a revised PPDAC approach is
> shown in the diagram below. As can be seen, although the clockwise
> sequence (1→5) applies as the principal flow, each stage may and often
> will feed back to the previous stage.

> In addition, it may well be beneficial to examine the process in the
> reverse direction, starting with Problem definition and then examining
> expectations as to the format and structure of the Conclusions
> (without pre-judging the outcomes!). This procedure then continues,
> step-by-step, in an anti-clockwise manner (e→a) determining the
> implications of these expectations for each stage of the process.

## Design of Experiments

Let us pretend we are Hamlet and design a simple experiment to find out
some "truth" (highly probable idea) that is important to us.

For example:

Let us consider the **Short Term Memory(STM)** of Foundation Students: 
(Or , we could use the **Guilford Divergent Thinking Test for Alternative Uses** as our vehicle for testing similar Hypotheses.)

### Hypothesis Design

What Hypothesis could be make? What do we have a hunch about?

-   STM scores are significantly different between students of Design
    and those of Art (;-O)
-  STM is much better for those who use Windows as compared to those who use Macs
-   Any other factors, not related to types of individuals, but more
    like parameters of STM itself, such as:
    -   Word complexity (Syllables per Word)
    -   No of Words
    -   Exposure Time and Recall Time
-   Students who speak their Mother Tongue fluently score better in
    Divergent Thinking Scores (Fluency, Flexibility, Elaboration,
    Originality)
-   There is a significant difference in Guilford Scores between
    students from "small towns" and those from metros.

### Factors in Hypothesis

In each of these Hypotheses, we can see that there are *binary* (Yes/No)
(two-valued) parameters that are part of the test: `Art vs Design`,` Windows vs Mac`, `Speaks Mother Tongue or Not`, `Small Town vs Metro`,
`Long vs Short List of Words`, `Short vs Long Words` and
`Long vs Short Exposure`.

If we want our experiment to be **fair** we must have:

-   Devise a test tool for each combination of binary parameters (*Test
    Type*)
-   the **same number** of subjects for each Test Type
-   **randomly** select the people as *respondents* each Test Type

We will choose one of these questions, or a similar one, and use the
**PPDAC Method** to proceed with our Experiment. This paper by Lawrence
to guide our Hypothesis Testing and the Design of our Experiment:

<u>[**Lawrence - A Design of Experiments Workshop as an Introduction to
Statistics - PDF**](Lawrence%20-%20A%20Design%20of%20Experiments%20Workshop%20as%20an%20Introduction%20to%20Statistics.pdf).</u>

This is the paper describing a simple *Design of Experiments* workshop
in class much like ours. We will try to mimic as much of this as we can.
Do read through this in the evening today in preparation for our
experiment.

### Data Analysis-1: Comparison of STM Histograms using Excel and WTFcsv

Once our test/survey is complete, let us collate all the data for all tests
using Excel. (This is for the STM Hypothesis; we can adapt this
spreadsheet for any other hypothesis test with categorical factors and
scores.)

1.  Download this <u>[**spreadsheet**](STM.xlsx)</u> and save as your own
    *named* local copy.
2.  Enter the data from all tests on the first sheet titled "RawData"
    (the one with the warning)
3.  **Duplicate this data** on the sheet titled "Data Duplicated". **Do
    not touch** the "RawData" sheet again. Good Practice!!
4.  For each of the *factors* under consideration, we will need to stack
    up STM scores from $2^3 = 8$ tests, i.e. eight columns.
5.  If we have three two-valued, categorical factors, we will get three
    sets of stacked scores, each from a (shuffled) set of the tests.
6.  Half the scores in these stacked columns will pertain to one *level*
    of the factor, and the other half of the scores will pertain to the
    other level. (This should remind you of the Karnaugh Map you may
    have learnt in your digital logic courses in school.). See example
    below: it shows how we can stack up the scores from *4* tests for
    each *level* of the parameter *p1*.

```{r,echo=FALSE,message=FALSE,warning=FALSE}

ggplot2::theme_set(new = theme_classic())

dat <- tibble(scores = rep(20, 8),
              test = LETTERS[1:8])
params <- expand_grid(p1 = c("True", "False"), p2 = c("Long", "Short"), p3 = c("Big", "Small"))

scores <- cbind(dat, params)
scores %>% kableExtra::kable(align = "c")

p1 <- ggplot(scores, aes(x = test, y = scores, fill = test)) + 
  geom_bar(stat = "identity", colour = "black") + 
  scale_fill_brewer(type = "seq",palette = "Set3") +
  theme_minimal() + 
  labs(title = "8 sets of 20 scores")

p2 <- ggplot(scores, aes(x = p1, y = scores, fill = test), colour = "black") + 
  geom_bar(stat = "identity", position = "stack", colour = "black") + 
  scale_fill_brewer(type = "seq",palette = "Set3") + 
  theme_minimal() + 
  labs(title = "Two sets of scores from 8 tests")

wrap_plots(p1,p2, ncol = 2, nrow = 1, guides = "collect")

```

1.  We will set up each of these *two stacked-score columns* in Col A on
    separate sheets, similar to the one shown in the sheet titled
    "Permutation Test". One sheet for each two-valued parameter.
2.  Create *paired* Histograms for each of the two stacked-score columns
    using WTFcsv, or Excel itself if you are confident. Ensure that
    there are *TWO* Histograms: one for each value of the factor under
    consideration.
3.  Inspection of the shapes and locations of these paired Histograms
    may give you an idea whether the factor under consideration has any
    effect on STM scores or none.

```{r,echo=FALSE,message=FALSE,warning=FALSE}
ggplot2::theme_set(new = theme_classic())

p1_T<- tibble(values = rpois(n = 80, lambda = 10),
              p1 = rep("HIGH", 80))
p1_F<- tibble(values = rpois(n = 80, lambda = 5),
              p1 = rep("LOW", 80))
p1 <- rbind(p1_T, p1_F)

ggplot(p1, aes(x = values, fill = p1, colour = p1)) + 
  geom_histogram(alpha = 0.3,) + theme_minimal() +
  geom_vline(xintercept = mean(p1$values[p1$p1 == "HIGH"]), colour = "red", linewidth = 2, linetype = "dotdash") + 
  geom_vline(xintercept = mean(p1$values[p1$p1 == "LOW"]), colour = "#00BFC4", linewidth = 2, linetype = "dotdash") +
  labs(title = "Paired STM score Histograms for Factor ***p1***",
      y = "Count of STM scores",
      x = "STM Scores"
  ) +
  geom_textbox(x = 2.5, y = 15, label = "Mean Score <br> for p1(LOW)", alpha = 0.1, width = unit(0.2, "npc"), fill = "cornsilk", colour = "#00BFC4") + 
  geom_textbox(x = 12.5, y = 12, label = "Mean Score <br>for p1(HIGH)", alpha = 0.1, width = unit(0.2, "npc"), fill = "cornsilk", colour = "red") + 
  theme(plot.title = ggtext::element_markdown())

```

4.  Save these 6 histograms as PNG files on your machines and use them
    along with *Comic Generator* (discussed below) to tell the story of
    your Hypothesis Testing.

### Computing Parallel Worlds: The Permutation Test

What is we cannot trust our eyes to compare the pair-wise Histograms?

If there was considerable overlap?

Is there a better way? Yes by playing another Childhood Game, with **a
deck of cards** !!!

#### Before we play with cards, some background

From: Tim C. Hesterberg (2015) *What Teachers Should Know About the
Bootstrap: Resampling in the Undergraduate Statistics Curriculum*, The
American Statistician, 69:4, 371-386, DOI: 10.1080/00031305.2015.1089789

> Student B. R. was annoyed by TV commercials. He suspected that there
> were more commercials in the "basic" TV channels, the ones that come
> with a cable TV subscription, than in the "extended" channels you pay
> extra for. To check this, he collected the data shown in Table 1. He
> measured an average of 9.21 minutes of commercials per half hour in
> the basic channels, vs only 6.87 minutes in the extended channels.
> This seems to support his hypothesis. But there is not much
> data---perhaps the difference was just random. The poor guy could only
> stand to watch 20 random half hours of TV. Actually, he didn't even do
> that---he got his girlfriend to watch half of it. (Are you as appalled
> by the deluge of commercials as I am? This is per half-hour!)

```{r echo = FALSE}
scores <- tibble(
  basic = c(6.95,10.013,10.62,10.15,8.583,7.62,8.233,10.35,11.016,8.516),
  extended = c(3.383,7.8,9.416,4.66,5.36,7.63,4.95,8.013,7.8,9.58))
scores %>% knitr::kable()
diff = mean(scores$basic) - mean(scores$extended)
diff

```

> The average difference in ad times between the two sets of TV channels
> is 2.34.

> How easy would it be for a difference of 2.34 minutes to occur just by
> chance? To answer this, we suppose there really is no difference
> between the two groups, that "basic" and "extended" are just labels.
> So what would happen if we assign labels randomly? How often would a
> difference like 2.34 occur?
>
> We'll pool all twenty observations, **randomly** pick 10 of them to
> label "basic" and label the rest "extended", and compute the
> difference in means between the two groups. We'll repeat that many
> times, say ten thousand, to get the *permutation distribution* shown
> in Figure 1.

```{r echo=FALSE, warning=FALSE, message=FALSE}
scores_long <- scores %>% pivot_longer(., cols = c("basic", "extended"), values_to = "times", names_to = "channel")
scores_long %>% knitr::kable()

null_dist_time <-
  do(10000) * diff(mean(
    scores_long$times ~ shuffle(scores_long$channel),
    data = scores_long))

null_dist_time %>% 
  ggplot(., aes(x = extended)) + 
  geom_histogram(aes(fill = extended >= 2.34)) +
  geom_vline(aes(xintercept = 2.34), colour = "green") + 
  labs(x = "Distribution of Diff in Ad Time Means under null hypothesis for Channel Type", title = "Figure1 - Is 2.34 possible by Chance?") + 
  guides(fill = guide_legend("Greater than 2.34"))

```

> The observed statistic 2.34 is also shown; the fraction of the
> distribution to the right of that value (≥ 2.34) is the probability
> that random labeling would give a difference that large. In this case,
> the probability of this value ( area coloured in green ) is \< 0.005.
>
> It would be rare for a difference this large to occur by chance. We
> have randomly tried "all possible chances" and are *hardly able to
> achieve similar,* and the rarer something is, the more likely that
> there is an underlying truth.
>
> And therefore we conclude there is a real difference between the
> groups and that ad time is different between `basic` and `extended` TV
> channels.

### Data Analysis-2: Creating Parallel Worlds

1.  We will execute on Permutation Test on one sheet step by step and
    decide whether that factor had a significant *effect* on STM scores.
2.  We **pretend** that there is no difference in scores whether the
    factor is chosen on way or another.
3.  We **mechanize this pretence** by lumping "both kinds" of scores
    together and shuffle them and divide them, **randomly** into two
    groups, and take the difference in scores.
4.  We can do this, like we did with our Monte Carlo experiment, many
    many times and calculate difference in scores each time. (It is like
    inventing many parallel worlds)
5.  If we look at the way these **randomly computed** scores are
    distributed and **compare** with the one measurement we did see, we
    can decide whether Mother Nature is up to something, or we are able
    to mimic the Mom.
6.  If we are **not** able to mimic Mom, then **Mom always knows** and
    we bow to her and ascribe a significance to the factor under
    consideration. Else, nope.
7.  Replicate this test for the other (binary) factors.

Apropos: how could we do this for non-binary factors ...??!!!

### Making a Data Comic out of our Experiment

What have been our Conclusions with the Experiment? Let us take our
Experiment and make a comic out of it:

**Data Comic Generator from Gramener**
[(**Weblink**)](https://gramener.com/comicgen)

<br> `r blogdown::shortcode("youtube", "E_2hdZuugI8")` <br>

### References

1.  Mahashweta Devi, *The Why Why Girl*,
    <u><https://www.tulikabooks.com/picture-books/the-why-why-girl-english.html></u>, read by Sarah Dryden Peterson
    <u><https://www.youtube.com/watch?v=xNcV5DZhg4c></u>

### Fun Stuff: Words Belong Hamlet

Hamlet's *To be or not to be* in Pidgin English <u>[Tok
Pisin](https://www.hawaii.edu/satocenter/langnet/definitions/tokpisin.html)</u>!

```{r echo=FALSE,eval=FALSE}
blogdown::shortcode("tweet", 
                    user = "bbcnewspidgin", 
                    id = "1108724221784530945")
                    
```


> Which way this time? Me killem de finish body b'long me. Or me no do
> 'im? Me no savvy.\
> Might 'e better 'long you-me catchem this fella, string for throw 'im
> this fella arrow.\
> Altogether b'long number one bad fella, name b'long him fortune? Me no
> savvy.\
> Might 'e better 'long you-me For fightem 'long altogether where him 'e
> makem you-me sorry too much.\
> Bimeby him fall down die finish? Me no savvy.
