---
title: "\U0001F350 Stat Tests for Correlations"
author: "Arvind Venkatadri"
date: Nov 22, 2022
lastmod: "`r Sys.Date()`"
weight: 20
type: book
output:
  html_document:
    toc_float: yes
    theme: flatly
    highlight: tango
    toc: yes
    code_folding: show
    code_download: yes
    number_sections: yes
df_print: paged
links:
- icon: circle
  icon_pack: fas
  name: Orange Tutorial
  url: "labs/Data-Analytics/correlations.ows"
- icon: sun
  icon_pack: fas
  name: Radiant
  url: "labs/Data-Analytics/correlations.rda"
- icon: r-project
  icon_pack: fab
  name: R Tutorial
  url: "labs/Data-Analytics/correlations.html"
- icon: database
  icon_pack: fas
  name: Sample Datasets
  url: "labs/Data-Analytics/data/qdd-data.zip"
keywords: Statistics ; Tests; p-value; Feynman Technique
abstract: This module is intended to assist with making statistically significant insights that drive business decisions. This document deals with the basics of stats. The method followed is that of Jonas Lindoloev, wherein every stat test is treated as a linear model `y = mx + c`.
editor_options:
  chunk_output_type: inline
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE, fig.align = "center", scipen = 1, digits = 3, fig.width = 7, fig.height = 5)
# CRAN Packages
library(tidyverse)
library(broom)
library(patchwork)
library(kableExtra)

library(mosaic)
library(mosaicCore)
library(mosaicData)

library(openintro) # datasets and methods
library(statsExpressions) # datasets and methods
library(ggstatsplot) # special stats plots
library(ggExtra)

# Non-CRAN Packages
# remotes::install_github("easystats/easystats")
library(easystats)

```

# Tests for Correlation

Correlation **r** is a measure of *strength* and *direction* of *linear
association* between two variables. **r** is between $[-1,+1]$, with $0$
implying no association/correlation.

From this definition, the *linear model* lends itself in a
straightforward way as a model to interpret *correlation*. Intuitively,
the slope of the linear model could be related to the correlation
between y and x.

## The Correlation Test

One of the basic Questions we would have of our data is: Does some
variable depend upon another in some way? Does $y$ depend upon $x$?

A **Correlation Test** is designed to answer exactly this question.

Let us now see how a Correlation Test can be re-formulated as a Linear
Model + Hypothesis Test.

Now we look at the numbers, using some toy data first and then a real
dataset.

```{r data_genr, echo=FALSE}
rnorm_fixed  <- function(N, mu = 0, sd = 1) {
  scale(rnorm(N))* sd + mu
}
```

```{r toy_data, echo=FALSE}
#set.seed(123456) # for replication
n = 500

# Data as vectors ( for t.tests etc)
x <- rnorm_fixed(n, mu = 0.0, sd = 1) #explanatory
y <- rnorm_fixed(n, mu = 0.3, sd = 2) # dependent #1
y2 <- rnorm_fixed(n, mu = 0.5, sd = 1.5) # dependent #2

# Make a tibble with all variables
mydata_wide <- tibble(x = x, y = y, y2 = y2)

# Long form data
mydata_long <- 
  mydata_wide %>%
  pivot_longer(., cols = c(x,y,y2), 
               names_to = "group", 
               values_to = "value")

# Long form data with only dependent variables
mydata_long_y <- 
  mydata_wide %>% 
  select(-x) %>% 
  pivot_longer(., cols = c(y,y2), 
               names_to = "group", 
               values_to = "value")
mydata_wide
mydata_long
mydata_long_y
```

## Pearson Correlation {.tabset .tabset-pills}

### Model

The model for Pearson Correlation tests is exactly the Linear Model:

$$
\begin{aligned}
y = \beta_0 + \beta_1 \times x\\
\\
H_0: Null\ Hypothesis\ => \beta_1 = 0\\\
H_a: Alternate\ Hypothesis\ => \beta_1 \ne 0\\
\end{aligned}
$$

### Code

```{r Pearson_Correlation}
#Pearson (0built-in test)
cor <- cor.test(y,x,method = "pearson") %>% 
  broom::tidy() %>% select(estimate, p.value)

#Linear Model
lin <- lm(y ~ 1 + x, data = mydata_wide) %>% 
  broom::tidy() %>% select(estimate, p.value)

# Scaled linear model
lin_scl <- lm(scale(y) ~ 1 + scale(x), data = mydata_wide) %>% 
  broom::tidy() %>% select(estimate, p.value)

print(cor)
print(lin)
print(lin_scl)
rbind(cor, lin, lin_scl) %>% print()
```

[[Notes:\\\\](Notes:\\){.uri}]([Notes:\\](Notes:\){.uri}){.uri} 1. The
*p-value* for Pearson Correlation and for the *slope* in the linear
model is the same ( 0.1053 ). Which means we cannot reject the NULL
hypothesis of "no relationship".

2.  Here is the relationship between the slope and correlation:

$$
Slope\ \beta_1 = \frac{sd_y}{sd_x} * r
$$

When both x and y have the same standard deviation, the slope and
correlation are the same. Here, since x has twice the `sd` of y, the
ratio of **slope** = `r lin$estimate[2]` to **r** = `r cor$estimate` is
0.5. Hence a linear model using `scale()` for both variables will show
slope = **r**.

Slope_Scaled: `r lin_scl$estimate[2]` = Correlation: `r cor$estimate`

Using the *linear model* method we get:

```{r}
# Linear Model
lin <- lm(y ~ 1 + x, data = mydata_wide) %>% 
  broom::tidy() %>% mutate(term = c("beta_0", "beta_1")) %>% select(term, estimate, p.value)
lin 

```

Why are $r$ and $\beta_1$ different, though the `p-value` is
suspiciously the same!?

Did we miss a factor of $\frac{-0.463}{-0.231} = 2$ somewhere...??

Let us **scale** the variables to within `{-1, +1}` : (subtract the mean
and divide by sd) and re-do the Linear Model with **scaled** versions
$x$ and $y$:

```{r}
# Scaled linear model
lin_scl <- lm(scale(y) ~ 1 + scale(x), data = mydata_wide) %>% 
  broom::tidy() %>% mutate(term = c("beta_0", "beta_1")) %>% select(term, estimate, p.value) %>% select(term, estimate, p.value)
lin_scl 

```

So we conclude:

1.  **When both x and y have the same standard deviation, the slope from
    the linear model and the Pearson correlation are the same**. Here,
    since x has twice the `sd` of y, the ratio of **slope** =
    `r lin$estimate[2]` to **r** = `r cor$estimate` is 0.5.

2.  There is this relationship between the **slope in the linear model**
    and **Pearson correlation**:

$$
Slope\ \beta_1 = \frac{sd_y}{sd_x} * r
$$

The slope is usually much more interpretable and informative than the
correlation coefficient.

2.  Hence a linear model using `scale()` for both variables will show
    slope = **r**.

Slope_Scaled: `r lin_scl$estimate[2]` = Correlation: `r cor$estimate`

3.  Finally, the *p-value* for Pearson Correlation and that for the
    *slope* in the linear model is the same ($0.1053$). Which means we
    cannot reject the NULL hypothesis of "no relationship".

### Example

TBD

## Spearman Correlation {.tabset .tabset-pills}

### Model

In some cases the **LINE** assumptions may not hold.

Nonlinear relationships, non-normally distributed data ( with large
**outliers** ) and working with *ordinal* rather than continuous data:
these situations necessitate the use of Spearman's *ranked* correlation
scores. (**Ranked**, not **sign-ranked**.).

See the example below: We choose to look at the `gpa_study_hours`
dataset. It has two numeric columns `gpa` and `study_hours`:

```{r gpa_study_hours}
glimpse(gpa_study_hours)

```

We can plot this:

```{r Pearson_example_3}

ggplot(gpa_study_hours, aes(x = study_hours, y = gpa)) + geom_point() + geom_smooth() + theme_minimal()
# ggstatsplot::ggscatterstats(data = gpa_study_hours, 
#                             x = study_hours, 
#                             y = gpa,
#                             marginal = TRUE,
#                             title = "GPA vs Study Hours")

```

Hmm...not normally distributed, and there is a sort of increasing
relationship, however is it linear? And there is some evidence of
heteroscedasticity, so the LINE assumptions are clearly in violation.
Pearson correlation would not be the best idea here.

Let us quickly try it anyway, using a Linear Model for the **scaled**
`gpa` and `study_hours` variables, from where we get:

```{r Pearson_example_1}


# Pearson Correlation as Linear Model
model_gpa <-
  lm(scale(gpa) ~ 1 + scale(study_hours), data = gpa_study_hours)

model_gpa %>%
  broom::tidy() %>% 
  mutate(term = c("beta_0", "beta_1")) %>% 
  cbind(confint(model_gpa) %>%                                              as_tibble()) %>%  
  select(term, estimate, p.value, `2.5 %`, `97.5 %`)

```

The correlation estimate is $0.133$; the `p-value` is $0.065$ (and the
`confidence interval` includes $0$).

Hence we fail to reject the NULL hypothesis that `study_hours` and `gpa`
have no relationship. But can this be right?

Should we use another test, that does not **need** the LINE assumptions?

### "Signed Rank" Values

Most statistical tests use the **actual values** of the data variables.
However, in some *non-parametric* statistical tests, the data are used
in **rank-transformed** sense/order. (In some cases the **signed-rank**
of the data values is used instead of the data itself.)

**Signed Rank** is calculated as follows:

1.  Take the absolute value of each observation in a sample

2.  Place the <u>*ranks*</u> in order of (absolute magnitude). The
    smallest number has *rank = 1* and so on. This gives is **ranked
    data**.

3.  Give each of the ranks the sign of the original observation ( + or
    -). This gives us **signed** ranked data.

```{r signed_rank_function}
signed_rank <- function(x) {sign(x) * rank(abs(x))}

```

Let us see how this might work by comparing data and its signed-rank
version...A quick set of plots:

```{r data_plots}
p1 <- ggplot(mydata_long,aes(x = group, y = value)) +
  geom_jitter(width = 0.02, height = 0,aes(colour = group), size = 4) +
  geom_segment(data = mydata_wide, aes(y = 0, yend = 0, 
                                       x = .75, 
                                       xend = 1.25 )) + 
  geom_text(aes(x = 1, y = 0.5, label = "0")) +
  geom_segment(data = mydata_wide, aes(y = 0.3, yend = 0.3, 
                                       x = 1.75 , 
                                       xend = 2.25 )) + 
  geom_text(aes(x = 2, y = 0.6, label = "0.3")) +
  geom_segment(data = mydata_wide, aes(y = 0.5, yend = 0.5, 
                                       x = 2.75, 
                                       xend = 3.25 )) + 
  geom_text(aes(x = 3, y = 0.8, label = "0.5")) +
  labs(title = "Original Data", subtitle = "Black Lines show Means") +
  ylab("Response Variable")

p2 <- mydata_long %>% 
  group_by(group) %>% 
  mutate( s_value = signed_rank(value)) %>% 
  ggplot(., aes(x = group, y = s_value)) + 
  geom_jitter(width = 0.02, height = 0,aes(colour = group), size = 4) + 
  stat_summary(fun = "mean", geom = "point", colour = "red", 
               size = 8) + 
  labs(title = "Signed Rank of Data", subtitle = "Red Points are means of Ranks!") +
  ylab("Signed Rank of Response Variable")

patchwork::wrap_plots(p1,p2, nrow = 1, guides = "collect")

```

So the means of the **ranks** three separate variables seem to be in the
same order as the means of the data variables themselves.

How about associations between data? Do ranks reflect well what the data
might?

```{r Spearman_Plot}
# Plot the data
p1 <- ggplot(mydata_wide, aes(x, y)) + 
  geom_point() +
  geom_smooth(method = "lm") +
  ggtitle(" Pearson Correlation\n and Linear Models")

# Plot ranked data
p2 <- mydata_wide %>% 
  mutate(x_rank = rank(x),
         y_rank = rank(y)) %>%
  ggplot(.,aes(x_rank, y_rank)) + 
  geom_point(shape = 15, size = 2) +
  geom_smooth(method = "lm") + 
  ggtitle(" Spearman Ranked Correlation\n and Linear Models")

patchwork::wrap_plots(p1,p2, nrow = 1, guides = "collect")

```

The slopes are almost identical, $0.25$ for both original data and
ranked data for $y\sim x$. So maybe *ranked* and even *sign_ranked* data
could work, and if it can work despite LINE assumptions not being
satisfied, that would be nice!

### How does Sign-Rank data work?

TBD: need to add some explanation here.

Spearman correlation = Pearson correlation using the rank of the data
observations. Let's check how this holds for a our x and y1 data:

So the Linear Model for the Ranked Data would be:

$$
\begin{aligned}
y = \beta_0 + \beta_1 \times rank(x)\\
\\
H_0: Null\ Hypothesis\ => \beta_1 = 0\\\
H_a: Alternate\ Hypothesis\ => \beta_1 \ne 0\\
\end{aligned}
$$

### Code

```{r Spearman_Correlation}
# Spearman
cor1 <- cor.test(x,y, method = "spearman") %>% 
  broom::tidy() %>% mutate(term = "Spearman Correlation ") %>% select(term, estimate, p.value)
cor1

```

```{r}
# Pearson using ranks
cor2 <- cor.test(rank(y), rank(x), method = "pearson") %>% 
broom::tidy() %>% select(estimate, p.value)
cor2
```

```{r}
# Linear Models using rank
cor3 <- lm(rank(y) ~ 1 + rank(x),data = mydata_wide) %>% 
  broom::tidy() %>% select(estimate, p.value)
cor3
```

Notes:

1.  When ranks are used, the slope of the linear model ($\beta_1$) has
    the same value as the Spearman correlation coefficient ( $\rho$ ).

2.  Note that the slope from the linear model now has an intuitive
    interpretation: **the number of ranks y changes for each change in
    rank of x**. ( Ranks are "independent" of `sd` )

### Example

We examine the `cars93` data, where the numeric variables of interest
are `weight` and `price`.

```{r Spearman_example_1}
cars93 %>% 
  ggplot(aes(weight, price)) + 
  geom_point() + geom_smooth(method = "lm", se = FALSE, lty = 2) + 
  labs(title = "Car Weight and Car Price have a nonlinear relationship")
```

Let us try a Spearman Correlation score for these variables, since the
data are not linearly related and the variance of `price` also is not
constant over `weight`

```{r Spearman_example_2}
cor.test(cars93$price, cars93$weight, method = "spearman") %>% broom::tidy()

# Using linear Model
lm(rank(price) ~ rank(weight), data = cars93) %>% summary()

# Stats Plot
ggstatsplot::ggscatterstats(data = cars93, x = weight, 
                            y = price,
                            type = "nonparametric",
                            title = "Cars93: Weight vs Price",
                            subtitle = "Spearman Correlation")
```

We see that using ranks of the `price` variable, we obtain a Spearman's
$\rho = 0.882$ with a `p-value` that is very small. Hence we are able to
reject the NULL hypothesis and state that there is a relationship
between these two variables. The **linear** relationship is evaluated as
a correlation of `0.882`.

```{r}
# Other ways using other packages
mosaic::cor_test(gpa ~ study_hours, data = gpa_study_hours) %>% broom:: tidy() %>% select(estimate, p.value, conf.low, conf.high)

```

```{r}

statsExpressions::corr_test(data = gpa_study_hours, 
                            x = study_hours, 
                            y = gpa)
```

# References

1.  *Common statistical tests are linear models (or: how to teach
    stats)* by [Jonas Kristoffer
    Lindeløv](https://lindeloev.github.io/tests-as-linear/)

2.  [CheatSheet](https://lindeloev.github.io/tests-as-linear/linear_tests_cheat_sheet.pdf)

3.  *Common statistical tests are linear models: a work through* by
    [Steve Doogue](https://steverxd.github.io/Stat_tests/)

4.  [Jeffrey Walker "Elements of Statistical Modeling for Experimental
    Biology"](https://www.middleprofessor.com/files/applied-biostatistics_bookdown/_book/)

5.  Diez, David M & Barr, Christopher D & Çetinkaya-Rundel, Mine:
    [OpenIntro Statistics](https://www.openintro.org/book/os/)

6.  Modern Statistics with R: From wrangling and exploring data to
    inference and predictive modelling by [Måns
    Thulin](http://www.modernstatisticswithr.com/)

7.  [Jeffrey Walker "A
    linear-model-can-be-fit-to-data-with-continuous-discrete-or-categorical-x-variables"](https://www.middleprofessor.com/files/applied-biostatistics_bookdown/_book/intro-linear-models.html#a-linear-model-can-be-fit-to-data-with-continuous-discrete-or-categorical-x-variables)
