---
title: "Sampling"
author: "Arvind Venkatadri"
date: '`r Sys.Date()`'
output:
  html_document:
    theme: flatly
    toc: TRUE
    toc_float: TRUE
    toc_depth: 2
    df_print: paged
    number_sections: TRUE
    code_folding: hide
    code_download: TRUE
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mosaic)
library(mosaicData)

library(NHANES)
library(cowplot)

```

# Introduction

Continuing to treat the `NHANES` dataset as a **population**, We will
try to replicate the process of sampling and CLT for another variable in
the NHANES variable, `AlcoholYear`.

## Summary for `AlcoholYear` population

```{r}

```

## Sampling `AlcoholYear`

```{r}

```

## Distribution and QQ Plot for the sample

```{r}

```

## Estimating Population Mean and Confidence Interval using the Sample

```{r}

```

## Case Study #2

Let us look at video game data from the University of Berkeley {Stats
Lab](

[Stat Labs Data Page](https://www.stat.berkeley.edu/users/statlabs/labs.html):

+------------+--------------------------------------------------------+
| Variable   |  Description                                           |
+============+========================================================+
|  time      |  Time spent playing video games in week prior to       |
|            | survey in hours.                                       |
+------------+--------------------------------------------------------+
|  like      |  Like to play video games\                             |
|            |  1=Never played, 2=Very much, 3=Somewhat, 4=Not        |
|            | really, 5=Not at all                                   |
+------------+--------------------------------------------------------+
|  where     |  Where play video games\                               |
|            |  1=Arcade, 2=Home on a system, 3=Home on a computer    |
|            | 4=Home on computer and system, 5=Arcade and            |
|            | Home(system or computer) 6=Arcade and home (both       |
|            | system and computer)                                   |
+------------+--------------------------------------------------------+
|  freq      |  How often play video games\                           |
|            |  1=Daily, 2=Weekly, 3=Monthly, 4=Semesterly            |
+------------+--------------------------------------------------------+
|  busy      |  Play if busy\                                         |
|            |   0=no, 1=yes                                          |
+------------+--------------------------------------------------------+
|  educ      |  Is playing video games educational?\                  |
|            |   0=no, 1=yes                                          |
+------------+--------------------------------------------------------+
|  sex       |  Sex\                                                  |
|            |   0=female, 1=male                                     |
+------------+--------------------------------------------------------+
|  age       |  Age in years                                          |
+------------+--------------------------------------------------------+
|  home      |  Computer at home\                                     |
|            |   0=no, 1=yes                                          |
+------------+--------------------------------------------------------+
|  math      |  Hate math\                                            |
|            |   0=no, 1=yes                                          |
+------------+--------------------------------------------------------+
|  work      |  Hours worked for pay in the week prior to survey      |
+------------+--------------------------------------------------------+
|  own       |  Own PC\                                               |
|            |   0=no, 1=yes                                          |
+------------+--------------------------------------------------------+
|  cdrom     |  Owned PC has a CDROM\                                 |
|            |   0=no, 1=yes                                          |
+------------+--------------------------------------------------------+
|  email     |  HAve email account\                                   |
|            |   0=no, 1=yes                                          |
+------------+--------------------------------------------------------+
|  grade     |  Grade expect in course\                               |
|            |   4=A, 3=B, 2=C, 1=D, 0=F                              |
+------------+--------------------------------------------------------+

## Sampling `time`

```{r}

```

## Distribution and QQ Plot for the sample

```{r}

```

## Estimating Population Mean and Confidence Interval using the Sample

```{r}

```



## Conclusion
Write up some comments here on your learnings on:\
- Sampling
- CLT



